# Cost burden and instability {#sec-burden}

<!-- Purpose: cost burden trends and housing instability — burden by tenure/race-ethnicity/
     income (ACS + CHAS + PUMS), wage affordability (OEWS), evictions, delinquency, and PIT
     homelessness counts. Theme-based takeaway H2s; static ggplot only; "Since 2022" callout
     per section; fig-alt throughout. Built in Task 11. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

burden_raw <- read_rds("data/acs_burden.rds")
chas       <- read_rds("data/chas.rds")
baseline   <- read_rds("data/baseline_2022.rds")

# Palette anchors — match ggtext subtitle spans to cb_pal fills
red    <- cb_pal[["Severely cost-burdened"]]
orange <- cb_pal[["Cost-burdened"]]
grey   <- cb_pal[["Not cost-burdened"]]

# Regional shares: sum estimates across the 8 rr localities (never average locality
# shares), excluding Virginia and the "Not computed" households from the denominator.
burden_reg <- burden_raw |>
  filter(geoid %in% pha, burden != "Not computed") |>
  summarise(estimate = sum(estimate), .by = c(year, tenure, burden)) |>
  mutate(share = estimate / sum(estimate), .by = c(year, tenure))

# Inline scalars for bullets and the Since-2022 callout
renter_burd_2024 <- burden_reg |>
  filter(year == 2024, tenure == "Renter", burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)

owner_burd_2024 <- burden_reg |>
  filter(year == 2024, tenure == "Owner", burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)

renter_sev_2024 <- burden_reg |>
  filter(year == 2024, tenure == "Renter", burden == "Severely cost-burdened") |>
  pull(share)

# Burdened counts and shares by tenure for 2014, 2019, and 2024 (regional totals)
burd_counts <- burden_reg |>
  filter(burden != "Not cost-burdened") |>
  summarise(estimate = sum(estimate), share = sum(share), .by = c(year, tenure))

renter_ct_2014 <- burd_counts |> filter(year == 2014, tenure == "Renter") |> pull(estimate)
renter_ct_2019 <- burd_counts |> filter(year == 2019, tenure == "Renter") |> pull(estimate)
renter_ct_2024 <- burd_counts |> filter(year == 2024, tenure == "Renter") |> pull(estimate)
owner_ct_2014  <- burd_counts |> filter(year == 2014, tenure == "Owner")  |> pull(estimate)
owner_ct_2019  <- burd_counts |> filter(year == 2019, tenure == "Owner")  |> pull(estimate)
owner_ct_2024  <- burd_counts |> filter(year == 2024, tenure == "Owner")  |> pull(estimate)
renter_sh_2019 <- burd_counts |> filter(year == 2019, tenure == "Renter") |> pull(share)
owner_sh_2019  <- burd_counts |> filter(year == 2019, tenure == "Owner")  |> pull(share)

# 2022 framework baselines (Fig 10.1): change in cost-burdened households since 2015
base_renter_chg <- baseline |>
  filter(metric == "cost_burdened_renters_change") |> pull(value)
base_owner_chg  <- baseline |>
  filter(metric == "cost_burdened_owners_change") |> pull(value)

# CHAS 2018-2022, pooled to the 8-locality region (sum estimates, never average
# locality shares — same rule as the ACS regional shares above)
chas_reg <- chas |>
  filter(geoid %in% pha) |>
  summarise(estimate = sum(estimate), .by = c(tenure, income_band, burden)) |>
  mutate(share = estimate / sum(estimate), .by = c(tenure, income_band))

chas_low_renter <- chas_reg |> filter(tenure == "Renter", income_band == "0-30% AMI",
                                       burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_low_owner <- chas_reg |> filter(tenure == "Owner", income_band == "0-30% AMI",
                                      burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_low_renter_sev <- chas_reg |> filter(tenure == "Renter", income_band == "0-30% AMI",
                                           burden == "Severely cost-burdened") |> pull(share)
chas_low_owner_sev <- chas_reg |> filter(tenure == "Owner", income_band == "0-30% AMI",
                                          burden == "Severely cost-burdened") |> pull(share)
chas_mid_renter <- chas_reg |> filter(tenure == "Renter", income_band == "50-80% AMI",
                                       burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_mid_owner <- chas_reg |> filter(tenure == "Owner", income_band == "50-80% AMI",
                                      burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_high_renter <- chas_reg |> filter(tenure == "Renter", income_band == "100%+ AMI",
                                        burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_high_owner <- chas_reg |> filter(tenure == "Owner", income_band == "100%+ AMI",
                                       burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
```

## Renter cost burden is climbing again since 2019

```{r}
#| label: fig-burden-trend
#| fig-cap: "Renter cost burden is climbing again since 2019"
#| fig-alt: "Two stacked bar charts of cost burden shares for renters and owners in the four primary Richmond region localities across the 2010-2014, 2015-2019, and 2020-2024 ACS periods, with 2010-2014 shown muted. Renter burden dips in 2015-2019 then rises in 2020-2024. Owner burden falls continuously over the full period."

trend_df <- burden_reg |>
  mutate(
    period = recode_values(year, 2014 ~ "2010–14", 2019 ~ "2015–19", 2024 ~ "2020–24"),
    burden = fct_rev(burden),
    period_alpha = if_else(year == 2014, 0.45, 1)
  )

ggplot(trend_df, aes(x = period, y = share, fill = burden, alpha = period_alpha)) +
  geom_col() +
  facet_wrap(~tenure) +
  scale_fill_manual(values = cb_pal) +
  scale_alpha_identity(guide = "none") +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Renter cost burden is climbing again since 2019",
    subtitle = paste0(
      "Share of households by cost burden, four primary localities<br>",
      "2010–14 shown muted for context<br>",
      "<span style='color:", orange, "'>**Cost-burdened**</span> = 30–49%; ",
      "<span style='color:", red, "'>**severely cost-burdened**</span> = 50%+"
    ),
    caption = acs_cap("B25070 & B25091", year = "2010-2014, 2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Renter cost burden eased to `r label_percent(accuracy = 1)(renter_sh_2019)` by 2015–19 but has since climbed back to `r label_percent(accuracy = 1)(renter_burd_2024)` in 2020–24 — a `r label_percent(accuracy = 1)(renter_burd_2024 - renter_sh_2019)` increase in just five years.**
- Severe burden drives the recent rise: severely cost-burdened renters grew from 24.5 percent of all renters in 2015–19 to `r label_percent(accuracy = 1)(renter_sev_2024)` in 2020–24.
- Owners kept moving the opposite direction: the burdened owner share fell from `r label_percent(accuracy = 1)(owner_sh_2019)` in 2015–19 to `r label_percent(accuracy = 1)(owner_burd_2024)` in 2020–24, continuing a decade-long decline from 26 percent in 2010–14.
- Because the renter population grew while the burdened share rose, the count of cost-burdened renter households climbed from `r label_comma()(renter_ct_2019)` in 2015–19 to `r label_comma()(renter_ct_2024)` in 2020–24 — a faster five-year rise than the change across the full prior decade.

::: {.callout-important}
## Since 2022
The 2022 Framework reported cost-burdened renters up by almost `r label_comma()(base_renter_chg)` households and cost-burdened owners down by more than `r label_comma()(abs(base_owner_chg))` since 2015 (Fig 10.1). This update finds the same split, larger on the renter side: burdened renter households up `r label_comma()(renter_ct_2024 - renter_ct_2014)` (+`r label_percent(accuracy = 0.1)((renter_ct_2024 - renter_ct_2014) / renter_ct_2014)`) and burdened owner households down `r label_comma()(abs(owner_ct_2024 - owner_ct_2014))` (`r label_percent(accuracy = 0.1)((owner_ct_2024 - owner_ct_2014) / owner_ct_2014)`) from 2010–2014 to 2020–2024. **Caveat:** the 2022 figures rode on the 2016–2020 ACS with a 2015 base year; this update compares non-overlapping 2010–2014 and 2020–2024 windows, so the magnitudes are directionally comparable but not measured on the same yardstick.
:::

## Renter cost burden tops 50 percent in Richmond, Henrico, and Chesterfield alike

```{r}
#| label: fig-burden-locality
#| fig-cap: "Renter cost burden tops 50 percent in Richmond, Henrico, and Chesterfield alike"
#| fig-alt: "Dodged bar chart of 2020-2024 renter cost burden shares for Virginia and the four primary localities. Each geography shows two bars: cost-burdened (30-49 percent of income) and severely cost-burdened (50 percent or more). Shares cluster between 22 and 28 percent for both categories everywhere; Richmond has the highest severe share at 28 percent, and Hanover is lowest overall."

loc_df <- burden_raw |>
  filter(geoid %in% c(pha, "51"), year == 2024, tenure == "Renter",
         burden %in% c("Cost-burdened", "Severely cost-burdened")) |>
  mutate(
    locality = str_remove(name, ", Virginia"),
    locality = fct_reorder(locality, share, .fun = sum) |> fct_relevel("Virginia")
  )

ggplot(loc_df, aes(x = locality, y = share, fill = burden)) +
  geom_col(position = position_dodge2(reverse = TRUE)) +
  coord_flip() +
  scale_fill_manual(values = cb_pal) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Renter cost burden tops 50 percent<br>in Richmond, Henrico, and Chesterfield alike",
    subtitle = paste0(
      "Share of renter households, 2020–2024<br>",
      "<span style='color:", orange, "'>**Cost-burdened**</span> (30–49% of income) and ",
      "<span style='color:", red, "'>**severely cost-burdened**</span> (50%+)"
    ),
    caption = acs_cap("B25070"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Total renter cost burden is nearly uniform across the three largest localities: 52 percent in Chesterfield, 52 percent in Richmond, and 52 percent in Henrico — this is a regional condition, not a city problem.**
- Richmond has the largest severe tilt: 28 percent of city renters pay half or more of their income toward housing, against 24 percent in Henrico and 25 percent in Chesterfield.
- Hanover is the outlier at 45 percent total renter burden, still well above the owner rate anywhere in the region.
- The three largest localities all exceed the statewide renter burden rate of 49 percent; Hanover sits below it.

## Cost burden is nearly universal below 30 percent of area median income

```{r}
#| label: fig-chas-ami
#| fig-cap: "Cost burden is nearly universal below 30 percent of area median income"
#| fig-alt: "Faceted bar chart of cost burden shares by household income band, as a percent of area median income (AMI), for owner and renter households in the four primary Richmond region localities. At 0-30% AMI, the large majority of renters and owners are cost-burdened, most of them severely. Burden falls steadily at each higher income band, dropping sharply above 100% AMI."

chas_plot <- chas_reg |>
  mutate(burden = fct_rev(burden))

ggplot(chas_plot, aes(x = income_band, y = share, fill = burden)) +
  geom_col() +
  facet_wrap(~tenure) +
  scale_fill_manual(values = cb_pal) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Cost burden is nearly universal below 30 percent of area median income",
    subtitle = paste0(
      "Share of households by cost burden and income band, four primary localities<br>",
      "<span style='color:", orange, "'>**Cost-burdened**</span> = 30–49% of income; ",
      "<span style='color:", red, "'>**severely cost-burdened**</span> = 50%+"
    ),
    caption = chas_cap("7"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none", axis.text.x = ggtext::element_markdown(angle = 30, hjust = 1)) +
  add_zero_line("y")
```

- **`r label_percent(accuracy = 0)(chas_low_renter)` of renters and `r label_percent(accuracy = 0)(chas_low_owner)` of owners below 30 percent of area median income are cost-burdened, and most of them severely: `r label_percent(accuracy = 0)(chas_low_renter_sev)` of renters and `r label_percent(accuracy = 0)(chas_low_owner_sev)` of owners in this band pay half or more of income toward housing.**
- Burden falls sharply as income rises: by 50-80% AMI, `r label_percent(accuracy = 0)(chas_mid_renter)` of renters and `r label_percent(accuracy = 0)(chas_mid_owner)` of owners are burdened; above 100% AMI, fewer than `r label_percent(accuracy = 0)(chas_high_renter)` of renters and `r label_percent(accuracy = 0)(chas_high_owner)` of owners are.
- Renters are more burdened than owners at every income band below 100% AMI, widest at the bottom of the scale.
- These are regional totals for the four primary localities.

## Severe burden is rising in the counties, easing in Richmond

```{r}
#| label: fig-severe-trend
#| fig-cap: "Severe burden is rising in the counties, easing in Richmond"
#| fig-alt: "Line chart of the severely cost-burdened renter share for the four primary localities across the 2010-2014, 2015-2019, and 2020-2024 ACS periods, with the earliest period muted. All four localities dip in 2015-2019, then diverge: Chesterfield rises from 21 to 25 percent between 2015-2019 and 2020-2024, Hanover from 19 to 22 percent, and Henrico from 22 to 24 percent. Richmond instead declines from 30 to 28 percent over the same recent span, though it remains highest overall."

sev_df <- burden_raw |>
  filter(geoid %in% pha, tenure == "Renter", burden == "Severely cost-burdened") |>
  mutate(
    locality = str_remove(name, " County, Virginia|, Virginia"),
    period = recode_values(year, 2014 ~ "2010–14", 2019 ~ "2015–19", 2024 ~ "2020–24"),
    is_recent = year >= 2019
  )

loc_pal <- c("Richmond city" = pha_pal[4], "Chesterfield" = pha_pal[6],
             "Henrico" = pha_pal[2], "Hanover" = pha_pal[1])

label_df <- sev_df |>
  filter(year == 2024) |>
  mutate(label_y = case_when(
    locality == "Chesterfield" ~ 0.262,
    locality == "Henrico"      ~ 0.237,
    locality == "Hanover"      ~ 0.212,
    .default = share
  ))

ggplot(sev_df, aes(x = period, y = share, color = locality, group = locality)) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = is_recent, alpha = is_recent)) +
  scale_size_manual(values = c(`TRUE` = 3, `FALSE` = 1.6), guide = "none") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.5), guide = "none") +
  ggtext::geom_richtext(
    data = label_df,
    aes(y = label_y, label = paste0("**", locality, "**")),
    hjust = 0, nudge_x = 0.08, size = 3.2, label.color = NA, fill = NA
  ) +
  scale_color_manual(values = loc_pal) +
  scale_y_continuous(labels = label_percent(accuracy = 1), limits = c(0, NA)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.35))) +
  labs(
    title = "Severe burden is rising in the counties, easing in Richmond",
    subtitle = paste0(
      "Share of renter households paying 50% or more of income on housing, primary localities<br>",
      "2010–14 shown muted for context · 2015–19 to 2020–24 change is the focus"
    ),
    caption = acs_cap("B25070", year = "2010-2014, 2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Severe renter burden jumped in every primary locality except Richmond between 2015–19 and 2020–24: Chesterfield rose 4.2 points to 25 percent, Hanover rose 2.7 points to 22 percent, and Henrico rose 1.9 points to 24 percent.**
- Richmond moved the other way, easing from 30 percent in 2015–19 to 28 percent in 2020–24 — even as it still carries the region's largest severe-burden caseload, about 15,800 renter households.
- All four localities had dipped in 2015–19 before this recent runup, so the full-decade change looks smaller than the last five years alone.
- Chesterfield added the most severely burdened renter households of any locality since 2015–19 — up about 1,600 (from 5,962 to 7,599), a 27 percent increase in five years.

## Planned content

::: {.callout-note title="Planned content"}
**Cost burden by race and ethnicity (CHAS).** HUD's CHAS tables will further break burden down by race and ethnicity, complementing the income-band cut above.
:::

::: {.callout-note title="To be added"}
**Evictions.** Eviction filings and judgments for the region's general district courts, trended against the 2022 Framework's pre-pandemic baseline.
:::

::: {.callout-note title="To be added"}
**Homelessness.** Greater Richmond CoC Point-in-Time counts, sheltered and unsheltered, updating the 2022 Framework's 2019–2021 series.
:::
