# Cost burden and instability {#sec-burden}

<!-- Purpose: cost burden trends and housing instability — burden by tenure/race-ethnicity/
     income (ACS + CHAS + PUMS), wage affordability (OEWS), evictions, delinquency, and PIT
     homelessness counts. Theme-based takeaway H2s; static ggplot only; "Since 2022" callout
     per section; fig-alt throughout. Built in Task 11. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

burden_raw <- read_rds("data/acs_burden.rds")
chas       <- read_rds("data/chas_cb.rds")
baseline   <- read_rds("data/baseline_2022.rds")

# Palette anchors — match ggtext subtitle spans to cb_pal fills
red    <- cb_pal[["Severely cost-burdened"]]
orange <- cb_pal[["Cost-burdened"]]
grey   <- cb_pal[["Not cost-burdened"]]

# Regional shares: sum estimates across the 8 rr localities (never average locality
# shares), excluding Virginia and the "Not computed" households from the denominator.
burden_reg <- burden_raw |>
  filter(geoid %in% pha, burden != "Not computed") |>
  summarise(estimate = sum(estimate), .by = c(year, tenure, burden)) |>
  mutate(share = estimate / sum(estimate), .by = c(year, tenure))

# Inline scalars for bullets and the Since-2022 callout
renter_burd_2024 <- burden_reg |>
  filter(year == 2024, tenure == "Renter", burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)

owner_burd_2024 <- burden_reg |>
  filter(year == 2024, tenure == "Owner", burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)

renter_sev_2024 <- burden_reg |>
  filter(year == 2024, tenure == "Renter", burden == "Severely cost-burdened") |>
  pull(share)

# Burdened counts and shares by tenure for 2014, 2019, and 2024 (regional totals)
burd_counts <- burden_reg |>
  filter(burden != "Not cost-burdened") |>
  summarise(estimate = sum(estimate), share = sum(share), .by = c(year, tenure))

renter_ct_2014 <- burd_counts |> filter(year == 2014, tenure == "Renter") |> pull(estimate)
renter_ct_2019 <- burd_counts |> filter(year == 2019, tenure == "Renter") |> pull(estimate)
renter_ct_2024 <- burd_counts |> filter(year == 2024, tenure == "Renter") |> pull(estimate)
owner_ct_2014  <- burd_counts |> filter(year == 2014, tenure == "Owner")  |> pull(estimate)
owner_ct_2019  <- burd_counts |> filter(year == 2019, tenure == "Owner")  |> pull(estimate)
owner_ct_2024  <- burd_counts |> filter(year == 2024, tenure == "Owner")  |> pull(estimate)
renter_sh_2019 <- burd_counts |> filter(year == 2019, tenure == "Renter") |> pull(share)
owner_sh_2019  <- burd_counts |> filter(year == 2019, tenure == "Owner")  |> pull(share)

# 2022 framework baselines (Fig 10.1): change in cost-burdened households since 2015
base_renter_chg <- baseline |>
  filter(metric == "cost_burdened_renters_change") |> pull(value)
base_owner_chg  <- baseline |>
  filter(metric == "cost_burdened_owners_change") |> pull(value)

# CHAS 2018-2022, pooled to the 8-locality region (sum estimates, never average
# locality shares — same rule as the ACS regional shares above)
chas_reg <- chas |>
  filter(geoid %in% pha) |>
  summarise(estimate = sum(estimate), .by = c(tenure, income_band, burden)) |>
  mutate(share = estimate / sum(estimate), .by = c(tenure, income_band))

chas_low_renter <- chas_reg |> filter(tenure == "Renter", income_band == "0-30% AMI",
                                       burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_low_owner <- chas_reg |> filter(tenure == "Owner", income_band == "0-30% AMI",
                                      burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_low_renter_sev <- chas_reg |> filter(tenure == "Renter", income_band == "0-30% AMI",
                                           burden == "Severely cost-burdened") |> pull(share)
chas_low_owner_sev <- chas_reg |> filter(tenure == "Owner", income_band == "0-30% AMI",
                                          burden == "Severely cost-burdened") |> pull(share)
chas_mid_renter <- chas_reg |> filter(tenure == "Renter", income_band == "50-80% AMI",
                                       burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_mid_owner <- chas_reg |> filter(tenure == "Owner", income_band == "50-80% AMI",
                                      burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_high_renter <- chas_reg |> filter(tenure == "Renter", income_band == "100%+ AMI",
                                        burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)
chas_high_owner <- chas_reg |> filter(tenure == "Owner", income_band == "100%+ AMI",
                                       burden != "Not cost-burdened") |>
  summarise(share = sum(share)) |> pull(share)

# Eviction filings, annual totals by locality (2016-2025; 2026 excluded as a partial year)
evictions_raw <- read_rds("data/evictions.rds")
evict_annual <- evictions_raw |>
  filter(year %in% 2016:2025) |>
  summarise(filings = sum(filings_count), .by = c(locality, year))
evict_reg <- evict_annual |> summarise(filings = sum(filings), .by = year)

evict_reg_2019 <- evict_reg |> filter(year == 2019) |> pull(filings)
evict_reg_2021 <- evict_reg |> filter(year == 2021) |> pull(filings)
evict_reg_2025 <- evict_reg |> filter(year == 2025) |> pull(filings)

evict_ches_2019 <- evict_annual |> filter(locality == "Chesterfield", year == 2019) |> pull(filings)
evict_ches_2025 <- evict_annual |> filter(locality == "Chesterfield", year == 2025) |> pull(filings)
evict_rich_2019 <- evict_annual |> filter(locality == "Richmond city", year == 2019) |> pull(filings)
evict_rich_2025 <- evict_annual |> filter(locality == "Richmond city", year == 2025) |> pull(filings)

base_evict_chg <- baseline |> filter(metric == "eviction_filings_pct_chg") |> pull(value)

# PIT homelessness counts, Greater Richmond CoC (2020-2025)
pit_raw <- read_rds("data/pit_homeless.rds")
pit_2025 <- pit_raw |> filter(year == 2025) |> pull(total_persons)
pit_2021 <- pit_raw |> filter(year == 2021) |> pull(total_persons)
pit_unsh_2024 <- pit_raw |> filter(year == 2024) |> pull(unsheltered)

base_pit_2019 <- baseline |> filter(metric == "pit_count", note == "2019") |> pull(value)
base_pit_2021 <- baseline |> filter(metric == "pit_count", str_detect(note, "^2021")) |> pull(value)

# VDOE McKinney-Vento homeless student counts (2020-21 through 2024-25)
vdoe_raw <- read_rds("data/vdoe_homeless.rds")
vdoe_reg <- vdoe_raw |> summarise(mv_count = sum(mv_count), .by = school_year)
vdoe_reg_2021 <- vdoe_reg |> filter(school_year == "2020-21") |> pull(mv_count)
vdoe_reg_2025 <- vdoe_reg |> filter(school_year == "2024-25") |> pull(mv_count)

vdoe_rich_2021 <- vdoe_raw |> filter(locality == "Richmond city", school_year == "2020-21") |> pull(mv_count)
vdoe_rich_2025 <- vdoe_raw |> filter(locality == "Richmond city", school_year == "2024-25") |> pull(mv_count)
vdoe_henrico_2021 <- vdoe_raw |> filter(locality == "Henrico", school_year == "2020-21") |> pull(mv_count)
vdoe_henrico_2025 <- vdoe_raw |> filter(locality == "Henrico", school_year == "2024-25") |> pull(mv_count)
vdoe_hanover_2025 <- vdoe_raw |> filter(locality == "Hanover", school_year == "2024-25") |> pull(mv_count)

base_student_chg <- baseline |> filter(metric == "student_homelessness_pct_chg") |> pull(value)

# Shared locality color scale, used by fig-severe-trend and the three trend charts below.
# Named lookups, not pha_pal[n] position, so the mapping survives a palette reorder.
loc_pal <- c("Richmond city" = unname(pha_color("Orange Deep")), "Chesterfield" = unname(pha_color("Navy")),
             "Henrico" = unname(pha_color("Blue")), "Hanover" = unname(pha_color("Teal")))
```

## Renter cost burden is climbing again since 2019

```{r}
#| label: fig-burden-trend
#| fig-cap: "Renter cost burden is climbing again since 2019"
#| fig-alt: "Two stacked bar charts of cost burden shares for renters and owners in the four primary Richmond region localities across the 2010-2014, 2015-2019, and 2020-2024 ACS periods, with 2010-2014 shown muted. Renter burden dips in 2015-2019 then rises in 2020-2024. Owner burden falls continuously over the full period."

trend_df <- burden_reg |>
  mutate(
    period = recode_values(year, 2014 ~ "2010–14", 2019 ~ "2015–19", 2024 ~ "2020–24"),
    burden = fct_rev(burden),
    period_alpha = if_else(year == 2014, 0.45, 1)
  )

ggplot(trend_df, aes(x = period, y = share, fill = burden, alpha = period_alpha)) +
  geom_col() +
  facet_wrap(~tenure) +
  scale_fill_manual(values = cb_pal) +
  scale_alpha_identity(guide = "none") +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Renter cost burden is climbing again since 2019",
    subtitle = paste0(
      "Share of households by cost burden, four primary localities<br>",
      "2010–14 shown muted for context<br>",
      "<span style='color:", orange, "'>**Cost-burdened**</span> = 30–49%; ",
      "<span style='color:", red, "'>**severely cost-burdened**</span> = 50%+"
    ),
    caption = acs_cap("B25070 & B25091", year = "2010-2014, 2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Renter cost burden eased to `r label_percent(accuracy = 1)(renter_sh_2019)` by 2015–19 but has since climbed back to `r label_percent(accuracy = 1)(renter_burd_2024)` in 2020–24 — a `r label_percent(accuracy = 1)(renter_burd_2024 - renter_sh_2019)` increase in just five years.**
- Severe burden drives the recent rise: severely cost-burdened renters grew from 24.5 percent of all renters in 2015–19 to `r label_percent(accuracy = 1)(renter_sev_2024)` in 2020–24.
- Owners kept moving the opposite direction: the burdened owner share fell from `r label_percent(accuracy = 1)(owner_sh_2019)` in 2015–19 to `r label_percent(accuracy = 1)(owner_burd_2024)` in 2020–24, continuing a decade-long decline from 26 percent in 2010–14.
- Because the renter population grew while the burdened share rose, the count of cost-burdened renter households climbed from `r label_comma()(renter_ct_2019)` in 2015–19 to `r label_comma()(renter_ct_2024)` in 2020–24 — a faster five-year rise than the change across the full prior decade.

::: {.callout-important}
## Since 2022
The 2022 Framework reported cost-burdened renters up by almost `r label_comma()(base_renter_chg)` households and cost-burdened owners down by more than `r label_comma()(abs(base_owner_chg))` since 2015 (Fig 10.1). This update finds the same split, larger on the renter side: burdened renter households up `r label_comma()(renter_ct_2024 - renter_ct_2014)` (+`r label_percent(accuracy = 0.1)((renter_ct_2024 - renter_ct_2014) / renter_ct_2014)`) and burdened owner households down `r label_comma()(abs(owner_ct_2024 - owner_ct_2014))` (`r label_percent(accuracy = 0.1)((owner_ct_2024 - owner_ct_2014) / owner_ct_2014)`) from 2010–2014 to 2020–2024. **Caveat:** the 2022 figures rode on the 2016–2020 ACS with a 2015 base year; this update compares non-overlapping 2010–2014 and 2020–2024 windows, so the magnitudes are directionally comparable but not measured on the same yardstick.
:::

## Renter cost burden tops 50 percent in Richmond, Henrico, and Chesterfield alike

```{r}
#| label: fig-burden-locality
#| fig-cap: "Renter cost burden tops 50 percent in Richmond, Henrico, and Chesterfield alike"
#| fig-alt: "Dodged bar chart of 2020-2024 renter cost burden shares for Virginia and the four primary localities. Each geography shows two bars: cost-burdened (30-49 percent of income) and severely cost-burdened (50 percent or more). Shares cluster between 22 and 28 percent for both categories everywhere; Richmond has the highest severe share at 28 percent, and Hanover is lowest overall."

loc_df <- burden_raw |>
  filter(geoid %in% c(pha, "51"), year == 2024, tenure == "Renter",
         burden %in% c("Cost-burdened", "Severely cost-burdened")) |>
  mutate(
    locality = str_remove(name, ", Virginia"),
    locality = fct_reorder(locality, share, .fun = sum) |> fct_relevel("Virginia")
  )

ggplot(loc_df, aes(x = locality, y = share, fill = burden)) +
  geom_col(position = position_dodge2(reverse = TRUE)) +
  coord_flip() +
  scale_fill_manual(values = cb_pal) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Renter cost burden tops 50 percent<br>in Richmond, Henrico, and Chesterfield alike",
    subtitle = paste0(
      "Share of renter households, 2020–2024<br>",
      "<span style='color:", orange, "'>**Cost-burdened**</span> (30–49% of income) and ",
      "<span style='color:", red, "'>**severely cost-burdened**</span> (50%+)"
    ),
    caption = acs_cap("B25070"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Total renter cost burden is nearly uniform across the three largest localities: 52 percent in Chesterfield, 52 percent in Richmond, and 52 percent in Henrico — this is a regional condition, not a city problem.**
- Richmond has the largest severe tilt: 28 percent of city renters pay half or more of their income toward housing, against 24 percent in Henrico and 25 percent in Chesterfield.
- Hanover is the outlier at 45 percent total renter burden, still well above the owner rate anywhere in the region.
- The three largest localities all exceed the statewide renter burden rate of 49 percent; Hanover sits below it.

## Cost burden is nearly universal below 30 percent of area median income

```{r}
#| label: fig-chas-ami
#| fig-cap: "Cost burden is nearly universal below 30 percent of area median income"
#| fig-alt: "Faceted bar chart of cost burden shares by household income band, as a percent of area median income (AMI), for owner and renter households in the four primary Richmond region localities. At 0-30% AMI, the large majority of renters and owners are cost-burdened, most of them severely. Burden falls steadily at each higher income band, dropping sharply above 100% AMI."

chas_plot <- chas_reg |>
  mutate(burden = fct_rev(burden))

ggplot(chas_plot, aes(x = income_band, y = share, fill = burden)) +
  geom_col() +
  facet_wrap(~tenure) +
  scale_fill_manual(values = cb_pal) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Cost burden is nearly universal below 30 percent of area median income",
    subtitle = paste0(
      "Share of households by cost burden and income band, four primary localities<br>",
      "<span style='color:", orange, "'>**Cost-burdened**</span> = 30–49% of income; ",
      "<span style='color:", red, "'>**severely cost-burdened**</span> = 50%+"
    ),
    caption = chas_cap("7"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none", axis.text.x = ggtext::element_markdown(angle = 30, hjust = 1)) +
  add_zero_line("y")
```

- **`r label_percent(accuracy = 0)(chas_low_renter)` of renters and `r label_percent(accuracy = 0)(chas_low_owner)` of owners below 30 percent of area median income are cost-burdened, and most of them severely: `r label_percent(accuracy = 0)(chas_low_renter_sev)` of renters and `r label_percent(accuracy = 0)(chas_low_owner_sev)` of owners in this band pay half or more of income toward housing.**
- Burden falls sharply as income rises: by 50-80% AMI, `r label_percent(accuracy = 0)(chas_mid_renter)` of renters and `r label_percent(accuracy = 0)(chas_mid_owner)` of owners are burdened; above 100% AMI, fewer than `r label_percent(accuracy = 0)(chas_high_renter)` of renters and `r label_percent(accuracy = 0)(chas_high_owner)` of owners are.
- Renters are more burdened than owners at every income band below 100% AMI, widest at the bottom of the scale.
- These are regional totals for the four primary localities.

## Severe burden is rising in the counties, easing in Richmond

```{r}
#| label: fig-severe-trend
#| fig-cap: "Severe burden is rising in the counties, easing in Richmond"
#| fig-alt: "Line chart of the severely cost-burdened renter share for the four primary localities across the 2010-2014, 2015-2019, and 2020-2024 ACS periods, with the earliest period muted. All four localities dip in 2015-2019, then diverge: Chesterfield rises from 21 to 25 percent between 2015-2019 and 2020-2024, Hanover from 19 to 22 percent, and Henrico from 22 to 24 percent. Richmond instead declines from 30 to 28 percent over the same recent span, though it remains highest overall."

sev_df <- burden_raw |>
  filter(geoid %in% pha, tenure == "Renter", burden == "Severely cost-burdened") |>
  mutate(
    locality = str_remove(name, " County, Virginia|, Virginia"),
    period = recode_values(year, 2014 ~ "2010–14", 2019 ~ "2015–19", 2024 ~ "2020–24"),
    is_recent = year >= 2019
  )

label_df <- sev_df |>
  filter(year == 2024) |>
  mutate(label_y = case_when(
    locality == "Chesterfield" ~ 0.262,
    locality == "Henrico"      ~ 0.237,
    locality == "Hanover"      ~ 0.212,
    .default = share
  ))

ggplot(sev_df, aes(x = period, y = share, color = locality, group = locality)) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = is_recent, alpha = is_recent)) +
  scale_size_manual(values = c(`TRUE` = 3, `FALSE` = 1.6), guide = "none") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.5), guide = "none") +
  ggtext::geom_richtext(
    data = label_df,
    aes(y = label_y, label = paste0("**", locality, "**")),
    hjust = 0, nudge_x = 0.08, size = 3.2, label.color = NA, fill = NA
  ) +
  scale_color_manual(values = loc_pal) +
  scale_y_continuous(labels = label_percent(accuracy = 1), limits = c(0, NA)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.35))) +
  labs(
    title = "Severe burden is rising in the counties, easing in Richmond",
    subtitle = paste0(
      "Share of renter households paying 50% or more of income on housing, primary localities<br>",
      "2010–14 shown muted for context · 2015–19 to 2020–24 change is the focus"
    ),
    caption = acs_cap("B25070", year = "2010-2014, 2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Severe renter burden jumped in every primary locality except Richmond between 2015–19 and 2020–24: Chesterfield rose 4.2 points to 25 percent, Hanover rose 2.7 points to 22 percent, and Henrico rose 1.9 points to 24 percent.**
- Richmond moved the other way, easing from 30 percent in 2015–19 to 28 percent in 2020–24 — even as it still carries the region's largest severe-burden caseload, about 15,800 renter households.
- All four localities had dipped in 2015–19 before this recent runup, so the full-decade change looks smaller than the last five years alone.
- Chesterfield added the most severely burdened renter households of any locality since 2015–19 — up about 1,600 (from 5,962 to 7,599), a 27 percent increase in five years.

## Eviction filings have nearly recovered regionally, but Richmond still files fewer than before the pandemic

```{r}
#| label: fig-eviction-trend
#| fig-cap: "Eviction filings have nearly recovered regionally, but Richmond still files fewer than before the pandemic"
#| fig-alt: "Line chart of annual eviction filings for the four primary localities, 2016 through 2025. All four localities drop sharply in 2020-2021, then recover. By 2025 Chesterfield filings exceed their 2019 level, Henrico and Hanover are near their 2019 level, and Richmond remains below its 2019 level."

ggplot(evict_annual, aes(x = year, y = filings, color = locality, group = locality)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(evict_annual, year == 2025), size = 2.5) +
  ggtext::geom_richtext(
    data = filter(evict_annual, year == 2025),
    aes(label = paste0("**", locality, "**")),
    hjust = 0, nudge_x = 0.2, size = 3.2, label.color = NA, fill = NA
  ) +
  scale_color_manual(values = loc_pal) +
  scale_x_continuous(breaks = seq(2016, 2025, 3), expand = expansion(mult = c(0.02, 0.2))) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Eviction filings have nearly recovered regionally,<br>but Richmond still files fewer than before the pandemic",
    subtitle = "Annual eviction filings, four primary localities, 2016–2025",
    caption = eviction_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Regional filings fell `r label_percent(accuracy = 1)(abs(evict_reg_2021 - evict_reg_2019) / evict_reg_2019)` from `r label_comma()(evict_reg_2019)` in 2019 to `r label_comma()(evict_reg_2021)` at the 2021 pandemic-era low, then rebounded to `r label_comma()(evict_reg_2025)` by 2025 — within `r label_percent(accuracy = 1)(abs(evict_reg_2025 - evict_reg_2019) / evict_reg_2019)` of the pre-pandemic total.**
- The recovery is uneven by locality: Chesterfield filings now exceed 2019, up from `r label_comma()(evict_ches_2019)` to `r label_comma()(evict_ches_2025)`, while Richmond remains below 2019, down from `r label_comma()(evict_rich_2019)` to `r label_comma()(evict_rich_2025)`.
- Henrico and Hanover filings are both within 1 percent of their 2019 totals.

::: {.callout-important}
## Since 2022
The 2022 Framework reported Richmond's average annual filings down `r abs(base_evict_chg)` percent by 2017–2019, a pre-pandemic decline. This update finds filings fell much further through the pandemic before rebounding; Richmond's 2025 total remains `r label_percent(accuracy = 1)(abs(evict_rich_2025 - evict_rich_2019) / evict_rich_2019)` below 2019.
:::

## Homelessness eased from its pandemic-era peak but remains above 2019 levels

```{r}
#| label: fig-pit-trend
#| fig-cap: "Homelessness eased from its pandemic-era peak but remains above 2019 levels"
#| fig-alt: "Stacked bar chart of sheltered and unsheltered persons counted in the Greater Richmond Continuum of Care's annual point-in-time count, 2020 through 2025. The total rises sharply from 546 in 2020 to a peak of 834 in 2021, then declines to 659 by 2025. Unsheltered persons trend upward across the period even as the sheltered total falls back."

pit_pal <- c("Sheltered" = unname(pha_color("Blue")), "Unsheltered" = unname(pha_color("Orange Deep")))

pit_plot <- pit_raw |>
  pivot_longer(c(sheltered, unsheltered), names_to = "status", values_to = "persons") |>
  mutate(status = str_to_title(status), status = fct_relevel(status, "Sheltered", "Unsheltered"))

ggplot(pit_plot, aes(x = factor(year), y = persons, fill = status)) +
  geom_col() +
  scale_fill_manual(values = pit_pal) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Homelessness eased from its pandemic-era peak<br>but remains above 2019 levels",
    subtitle = paste0(
      "Point-in-time count of persons experiencing homelessness, Greater Richmond CoC<br>",
      "<span style='color:", pit_pal[["Sheltered"]], "'>**Sheltered**</span> and ",
      "<span style='color:", pit_pal[["Unsheltered"]], "'>**unsheltered**</span>"
    ),
    caption = pit_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **The point-in-time count fell to `r label_comma()(pit_2025)` in 2025, down `r label_percent(accuracy = 1)(abs(pit_2025 - pit_2021) / pit_2021)` from the 2021 peak of `r label_comma()(pit_2021)`, but still `r label_percent(accuracy = 1)((pit_2025 - base_pit_2019) / base_pit_2019)` above the 2019 count of `r label_comma()(base_pit_2019)`.**
- Unsheltered homelessness trended upward even as the total eased, peaking at `r label_comma()(pit_unsh_2024)` persons in 2024.
- Sheltered persons made up the large majority of the count in every year, ranging from about 70 to 88 percent of the total.

::: {.callout-important}
## Since 2022
The 2022 Framework flagged the COVID-era jump in the point-in-time count from `r label_comma()(base_pit_2019)` in 2019 to `r label_comma()(base_pit_2021)` in 2021 (+68 percent). This update finds the count has since eased but not returned to baseline: `r label_comma()(pit_2025)` in 2025 remains `r label_percent(accuracy = 1)((pit_2025 - base_pit_2019) / base_pit_2019)` above 2019.
:::

## Homeless student counts have nearly doubled regionally since the pandemic

```{r}
#| label: fig-vdoe-trend
#| fig-cap: "Homeless student counts have nearly doubled regionally since the pandemic"
#| fig-alt: "Line chart of McKinney-Vento identified homeless student counts for the four primary localities' school divisions, 2020-21 through 2024-25. All four localities rise over the period. Henrico grows fastest, more than doubling from 551 to 1,219 students; Richmond, Chesterfield, and Hanover each grow at a slower but still substantial pace."

ggplot(vdoe_raw, aes(x = school_year, y = mv_count, color = locality, group = locality)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(vdoe_raw, school_year == "2024-25"), size = 2.5) +
  ggtext::geom_richtext(
    data = filter(vdoe_raw, school_year == "2024-25"),
    aes(label = paste0("**", locality, "**")),
    hjust = 0, nudge_x = 0.1, size = 3.2, label.color = NA, fill = NA
  ) +
  scale_color_manual(values = loc_pal) +
  scale_x_discrete(expand = expansion(mult = c(0.02, 0.22))) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Homeless student counts have nearly doubled<br>regionally since the pandemic",
    subtitle = "McKinney-Vento identified homeless students, four primary localities' school divisions",
    caption = vdoe_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none") +
  add_zero_line("y")
```

- **Henrico's count more than doubled, from `r label_comma()(vdoe_henrico_2021)` in 2020-21 to `r label_comma()(vdoe_henrico_2025)` in 2024-25 (+`r label_percent(accuracy = 1)((vdoe_henrico_2025 - vdoe_henrico_2021) / vdoe_henrico_2021)`), the largest increase in the region.**
- Richmond's count rose `r label_percent(accuracy = 1)((vdoe_rich_2025 - vdoe_rich_2021) / vdoe_rich_2021)` over the same span, from `r label_comma()(vdoe_rich_2021)` to `r label_comma()(vdoe_rich_2025)` students.
- Regional totals rose from `r label_comma()(vdoe_reg_2021)` to `r label_comma()(vdoe_reg_2025)` identified homeless students — a `r label_percent(accuracy = 1)((vdoe_reg_2025 - vdoe_reg_2021) / vdoe_reg_2021)` increase in five school years.
- Hanover remains the smallest count in the region at `r label_comma()(vdoe_hanover_2025)` students in 2024-25, though it grew at a similar rate to Chesterfield and Richmond.

::: {.callout-important}
## Since 2022
The 2022 Framework found Richmond Public Schools' homeless student count had fallen `r abs(base_student_chg)` percent from 2017-18 to 2019-20, a pre-pandemic decline. This update finds the opposite direction since the pandemic: Richmond's count climbed `r label_percent(accuracy = 1)((vdoe_rich_2025 - vdoe_rich_2021) / vdoe_rich_2021)` from 2020-21 to 2024-25, reaching `r label_comma()(vdoe_rich_2025)` students. **Caveat:** the two periods are not directly comparable — the 2022 figure covers a pre-pandemic two-year window, while this update covers five post-pandemic school years.
:::

## Planned content

::: {.callout-note title="Planned content"}
**Cost burden by race and ethnicity (CHAS).** HUD's CHAS tables will further break burden down by race and ethnicity, complementing the income-band cut above.
:::
