# Housing demand {#sec-demand}

<!-- Purpose: who lives in the region and how demand is changing — population, components
     of change, households by type/size/age, projections. Theme-based takeaway H2s; static
     ggplot only; "Since 2022" callout; fig-alt throughout. Narrative rule: takeaway titles
     + bullet findings + callouts only — no drafted prose. Built in Task 7. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

decennial <- read_rds("data/decennial.rds")
pep       <- read_rds("data/pep.rds")
wcoop     <- read_rds("data/wcoop.rds")
demog     <- read_rds("data/acs_demographics.rds")
income    <- read_rds("data/acs_income.rds")
baseline  <- read_rds("data/baseline_2022.rds")

# Palette anchors — match ggtext subtitle spans to fills. Named lookups, not
# pha_pal[n] position, so the mapping survives a palette reorder.
green <- unname(pha_color("Teal")); lblue <- unname(pha_color("Blue"))
orange <- unname(pha_color("Orange")); dblue <- unname(pha_color("Navy"))

# Inline scalars — Since-2022 callout (baseline projections cover the 4-locality region)
pha4 <- unname(pha)

base_proj_2050 <- baseline |>
  filter(metric == "population_projection_2050", geography == "region") |> pull(value)
base_proj_pct <- baseline |>
  filter(metric == "population_projection_pct_chg", geography == "region") |> pull(value)

pop_2020_4loc <- decennial |>
  filter(year == 2020, label == "Total population", geoid %in% pha4) |>
  summarise(sum(value)) |> pull()
proj_2050_4loc <- wcoop |>
  filter(series == "total_population", year == 2050, geoid %in% pha4) |>
  summarise(sum(estimate)) |> pull()
cur_proj_pct <- proj_2050_4loc / pop_2020_4loc - 1
```

::: {.callout-note}
## Since 2022

The 2022 Data Update projected the primary region would reach
**`r label_comma()(base_proj_2050)` residents by 2050**, growth of
**+`r base_proj_pct`%** from 2020. The Weldon Cooper Center's updated 2025-release
projections now put the 2050 total at **`r label_comma()(round(proj_2050_4loc))`**
(**`r label_percent(accuracy = 0.1)(cur_proj_pct)`** growth), a
`r label_percent(accuracy = 0.1)(proj_2050_4loc / base_proj_2050 - 1)` upward revision
of the 2050 figure. The region's growth outlook has strengthened, not softened, since
the Framework was published.
:::

## The region's population continues to grow

```{r}
#| label: fig-pop-trend
#| fig-cap: "Over 50,000 new residents since the 2020 census"
#| fig-alt: "Line chart of total population in the four primary Richmond region localities. Annual population estimates from 2020 to 2025 are highlighted in teal, showing continued growth. Decennial census counts for 2000, 2010, and 2020 are shown muted in the background for long-term context."

pop_hist <- decennial |>
  filter(label == "Total population", geoid %in% pha) |>
  summarise(pop = sum(value), .by = year)

pop_recent <- pep |>
  filter(geoid %in% pha, variable == "POPESTIMATE") |>
  summarise(pop = sum(value), .by = year)

pop_latest <- pop_recent |> slice_max(year, n = 1)
pop_chg <- pop_latest$pop - pop_2020_4loc
pop_pct <- pop_chg / pop_2020_4loc

ggplot() +
  geom_line(data = pop_hist, aes(year, pop), color = dblue, linewidth = 0.6, alpha = 0.45) +
  geom_point(data = pop_hist, aes(year, pop), color = dblue, size = 1.8, alpha = 0.45) +
  geom_line(data = pop_recent, aes(year, pop), color = green, linewidth = 1.4) +
  geom_point(data = pop_recent, aes(year, pop), color = green, size = 3) +
  geom_text(
    data = pop_latest, aes(year, pop, label = label_comma()(pop)),
    color = green, fontface = "bold", hjust = 0, nudge_x = 0.4, size = 3.4
  ) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  scale_x_continuous(breaks = c(2000, 2010, 2020, 2025), limits = c(2000, 2027.5)) +
  add_zero_line() +
  labs(
    title = "Over 50,000 new residents since the 2020 census",
    subtitle = paste0(
      "<span style='color:", dblue, "'>**2000-2020 census counts**</span> shown for context, ",
      "<span style='color:", green, "'>**2020-2025 annual estimates**</span> highlighted"
    ),
    caption = paste(pep_cap(), "Decennial Census, 2000-2020."),
    x = NULL, y = NULL
  ) +
  theme_pha()
```

- **The core region added about `r label_comma()(round(pop_chg, -3))` residents between the 2020 census and the 2025 population estimate, growth of `r label_percent(accuracy = 0.1)(pop_pct)` in five years.**
- Every year from 2020 to 2025 added population, with the largest single-year gains in 2024 and 2025.

## Most growth still comes from people moving into region

```{r}
#| label: fig-components
#| fig-cap: "Migration, not births, drives the region's population growth"
#| fig-alt: "Grouped bar chart of annual components of population change in the four primary localities, 2021 to 2025. Net migration bars are substantially larger than natural change bars in every year. Natural change (births minus deaths) stays small — well under 2,000 residents per year."

components <- pep |>
  filter(geoid %in% pha, variable %in% c("NATURALCHG", "NETMIG"), year >= 2021) |>
  summarise(value = sum(value), .by = c(year, label))

mig_peak <- components |> filter(label == "Net migration") |> slice_max(value, n = 1)

ggplot(components, aes(factor(year), value, fill = label)) +
  geom_col(position = position_dodge2(padding = 0.1)) +
  scale_fill_manual(values = c("Natural change" = dblue, "Net migration" = green)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Migration, not births, drives the region's population growth",
    subtitle = paste0(
      "Annual components of population change, 2021-2025<br>",
      "<span style='color:", green, "'>**Net migration**</span> vs. ",
      "<span style='color:", dblue, "'>**natural change** (births minus deaths)</span>"
    ),
    caption = pep_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **Net migration has delivered the large majority of the region's growth every year since 2021, peaking at roughly `r label_comma()(round(mig_peak$value, -2))` net movers in `r mig_peak$year`.**
- Natural change (births minus deaths) has stayed small — under about 2,100 residents per year regionwide — reflecting an aging population.
- Overall, the region's demand pressure is *primarily* driven by households arriving from elsewhere, who need housing upon arrival.
- The 2020 estimate period covers only April-June 2020 and is omitted from the chart.

## Every locality is projected to keep growing

```{r}
#| label: fig-projections
#| fig-cap: "Chesterfield is projected to pass 500,000 residents by 2050"
#| fig-alt: "Line chart of projected total population for the four primary localities from the 2020 census to 2050. Chesterfield rises from about 365,000 to 520,000, Henrico from 334,000 to 416,000, Richmond city from 227,000 to 276,000, and Hanover from 110,000 to 139,000. All four lines rise across the whole period."

proj <- bind_rows(
  decennial |>
    filter(year == 2020, label == "Total population", geoid %in% pha4) |>
    select(geoid, name, year, pop = value),
  wcoop |>
    filter(series == "total_population", geoid %in% pha4) |>
    select(geoid, name, year, pop = estimate)
) |>
  mutate(locality = str_remove(name, ", Virginia$"))

proj_labels <- proj |> filter(year == 2050)

ggplot(proj, aes(year, pop, color = locality)) +
  geom_line(linewidth = 1, linetype = "42") +
  geom_point(size = 2.5) +
  geom_text(
    data = proj_labels,
    aes(label = paste0(str_remove(locality, " County"), "\n", label_comma()(round(pop)))),
    hjust = 0, nudge_x = 1.5, size = 3.2, fontface = "bold", lineheight = 0.9,
    show.legend = FALSE
  ) +
  scale_color_manual(values = unname(pha_colors[c("Orange", "Teal", "Blue", "Navy")])) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  scale_x_continuous(breaks = c(2020, 2030, 2040, 2050), limits = c(2020, 2058)) +
  labs(
    title = "Chesterfield is projected to pass 500,000 residents by 2050",
    subtitle = "Projected total population, four primary localities<br>2020 census count to 2050",
    caption = paste(dec_cap("2020"), str_remove(wc_cap(), fixed("**Source:** "))),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none")
```

- **Chesterfield is projected to grow 43% by 2050, reaching about 520,000 residents — up from the 38% growth projected in the 2022 Framework.**
- Henrico (+25%, to 416,000), Hanover (+27%, to 139,000), and Richmond city (+22%, to 276,000) all continue growing, with no primary locality projected to plateau.

## Older adults becoming larger share of population

```{r}
#| label: fig-age
#| fig-cap: "Nearly one in three residents is now 55 or older"
#| fig-alt: "Population pyramid of the four primary localities by age group and sex from the 2020-2024 ACS. The 25-34 age group is the largest. Residents 55 and older account for nearly a third of the population. Male and female bars are roughly symmetric, with women outnumbering men at ages 75 and over."

age_groups <- c(
  "Under 5 years" = "Under 5", "5 to 9 years" = "5-14", "10 to 14 years" = "5-14",
  "15 to 17 years" = "15-24", "18 and 19 years" = "15-24", "20 years" = "15-24",
  "21 years" = "15-24", "22 to 24 years" = "15-24",
  "25 to 29 years" = "25-34", "30 to 34 years" = "25-34",
  "35 to 39 years" = "35-44", "40 to 44 years" = "35-44",
  "45 to 49 years" = "45-54", "50 to 54 years" = "45-54",
  "55 to 59 years" = "55-64", "60 and 61 years" = "55-64", "62 to 64 years" = "55-64",
  "65 and 66 years" = "65-74", "67 to 69 years" = "65-74", "70 to 74 years" = "65-74",
  "75 to 79 years" = "75+", "80 to 84 years" = "75+", "85 years and over" = "75+"
)

age_sex <- demog |>
  filter(table == "B01001", geoid %in% pha) |>
  separate_wider_delim(label, "!!", names = c("total", "sex", "age"),
                       too_few = "align_start") |>
  filter(!is.na(age)) |>
  mutate(
    sex = str_remove(sex, ":$"),
    age_group = factor(age_groups[age], levels = unique(age_groups))
  ) |>
  summarise(pop = sum(estimate), .by = c(sex, age_group))

ggplot(age_sex, aes(x = if_else(sex == "Male", -pop, pop), y = age_group, fill = sex)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = c("Male" = dblue, "Female" = green)) +
  scale_x_continuous(labels = \(x) label_comma()(abs(x))) +
  labs(
    title = "Nearly one in three residents is now 55 or older",
    subtitle = paste0(
      "Population by age group, 2020-2024 average<br>",
      "<span style='color:", dblue, "'>**Male**</span> and ",
      "<span style='color:", green, "'>**female**</span>"
    ),
    caption = acs_cap("B01001"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("x")
```

- **Residents 55 and older make up 30% of the region's population, and residents 65 and older make up 17%.**
- The 25-34 age group is the region's single largest — prime household-formation years, which converts directly into near-term housing demand.
- Seniors 65+ living alone represent a meaningful share of all households in the four primary localities — a group with distinct needs for smaller, accessible, and lower-maintenance homes.
- Women outnumber men in the 75+ group, so older single-person households are disproportionately older women.

## Median household incomes continue to rise

```{r}
#| label: fig-income
#| fig-cap: "Households in every locality have seen incomes grow"
#| fig-alt: "Dot plot of median household income for the four primary localities at three ACS vintages: 2014, 2019, and 2024. The 2014 point is muted grey; the 2019-to-2024 segment is highlighted. Richmond city is lowest in 2024 at about $65,000. All four localities show consistent income gains since both 2014 and 2019."

mhi <- income |>
  filter(table == "B19013", geoid %in% pha) |>
  mutate(
    locality = str_remove(name, ", Virginia$") |> str_remove(" County"),
    locality = fct_reorder(locality, estimate, .fun = max)
  )

mhi_wide <- mhi |>
  select(locality, year, estimate) |>
  pivot_wider(names_from = year, values_from = estimate, names_prefix = "y")

ggplot(mhi, aes(estimate, locality)) +
  geom_segment(
    data = mhi_wide, aes(x = y2014, xend = y2019, y = locality, yend = locality),
    color = "grey85", linewidth = 0.8
  ) +
  geom_segment(
    data = mhi_wide, aes(x = y2019, xend = y2024, y = locality, yend = locality),
    color = green, linewidth = 1.3
  ) +
  geom_point(aes(color = factor(year), size = factor(year))) +
  scale_color_manual(values = c("2014" = "grey65", "2019" = orange, "2024" = dblue)) +
  scale_size_manual(values = c("2014" = 2, "2019" = 2.8, "2024" = 3.2), guide = "none") +
  scale_x_continuous(labels = label_dollar()) +
  labs(
    title = "Households in every locality have seen incomes grow",
    subtitle = paste0(
      "Median household income by locality, nominal dollars<br>",
      "<span style='color:", orange, "'>**2019**</span> to ",
      "<span style='color:", dblue, "'>**2024**</span> change highlighted; ",
      "2014 shown muted for context"
    ),
    caption = acs_cap("B19013", "2010-2014, 2015-2019, and 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  theme(legend.position = "none")
```

- **Every locality's median household income rose at least 23 percent since 2019.**
- Richmond city's median rose 37 percent since 2019 — from $47,250 to $64,587 — yet it remains the lowest among the four primary localities.
- The slowest income growth among the four primary localities since 2019 was in Chesterfield (+23%).
- Going back to 2014, every primary locality's median has risen at least 40% in nominal terms — a full decade of income growth that predates this recent acceleration.
- Each vintage is expressed in its own survey-year dollars, so these changes are nominal; part of the increase reflects inflation.

::: {.callout-note}
## Data caveats

- Population estimates for 2020-2025 are Census Bureau Vintage 2025 figures and will be revised in future vintages. Weldon Cooper projections are the 2025 release, benchmarked to the 2020 census.
- ACS income medians are 5-year averages in each vintage's own survey-year dollars. Comparisons across vintages are nominal, not inflation-adjusted.
- The 2022 Data Update baseline used 2016-2020 ACS and the prior Weldon Cooper release, so "Since 2022" comparisons reflect both real change and vintage revisions.
:::
