# Housing gap {#sec-gaps}

<!-- Purpose: the affordable-and-available rental gap by AMI (PUMS, NLIHC-style), the
     starter-home gap (renters who could buy), and minimum income for homeownership by
     price/size/type vs household income profiles. Theme-based takeaway H2s; static ggplot
     only; "Since 2022" callout per section; fig-alt throughout.
     Sections 3-4 built from MLS + FRED + ACS + CHAS + HUD income limits (r/gaps.R,
     r/affordcalc.R, r/hud_ami.R) -- a preliminary draft ahead of the PUMS pipeline
     (Section 5 row 15), which still gates sections 1-2. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

income_table  <- read_rds("data/gaps_income_table.rds")
starter       <- read_rds("data/gaps_starter.rds")
baseline      <- read_rds("data/baseline_2022.rds")

# Named lookups, not pha_pal[n] position, so the mapping survives a palette reorder.
green <- unname(pha_color("Teal")); orange <- unname(pha_color("Orange"))
red <- unname(pha_color("Orange Deep")); dblue <- unname(pha_color("Navy"))

entry_all <- income_table |> filter(structure_type == "All types", price_tier == "Entry-level (25th pct.)")
median_all <- income_table |> filter(structure_type == "All types", price_tier == "Typical (median)")

st <- starter$summary
bands <- starter$renter_bands

base_140k <- baseline |> filter(metric == "max_affordable_home_price_by_wage") |> pull(value)
base_gap80 <- baseline |> filter(metric == "rental_gap_80ami") |> slice(1) |> pull(value)
base_gap30 <- baseline |> filter(metric == "rental_gap_30ami") |> pull(value)
```

This section quantifies the region's housing shortfall three ways: rental units affordable and available by income band, renters positioned to become owners, and the income required to buy at today's prices.

## Affordable-and-available rental gap by income

::: {.callout-note title="To be added"}
An NLIHC-style bar chart of the surplus or deficit of rental units that are both affordable to and available to each AMI band (30/50/80/100/120% AMI), for the region's core three localities (Chesterfield, Henrico, Richmond city). Built from detailed Census microdata, with AMI bands assigned from HUD's FY2026 Income Limits by household size.
:::

## Competition for limited supply of affordable rentals

::: {.callout-note title="To be added"}
A table of renter households versus affordable units versus affordable-and-available units by AMI band, showing how higher-income occupancy of low-cost units compounds the shortage at the bottom. Source: detailed Census microdata with HUD's FY2026 Income Limits.
:::

## Starter homes out of reach for most renters

```{r}
#| label: fig-starter-gap
#| fig-cap: "Most renter households fall short of the income a starter home now requires"
#| fig-alt: "Column chart of renter households in the four primary localities by HUD AMI band, from HUD CHAS 2018-2022 data. Households in the 0-30, 30-50, and 50-80 percent AMI bands are shaded dark orange, marking incomes below what buying the region's entry-level home requires. The 80-100 percent AMI band is shaded orange, since the entry-level threshold falls inside that band. The 100 percent and above band is shaded teal. Band sizes are roughly 30,500, 24,300, 33,200, 16,700, and 35,000 households."

bands_plot <- bands |>
  mutate(status = case_when(
    income_band %in% c("0-30% AMI", "30-50% AMI", "50-80% AMI") ~ "Below the entry-level threshold",
    income_band == "80-100% AMI" ~ "Straddles the entry-level threshold",
    income_band == "100%+ AMI" ~ "At or above the entry-level threshold"
  ))

ggplot(bands_plot, aes(x = income_band, y = households, fill = status)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c(
    "Below the entry-level threshold" = red,
    "Straddles the entry-level threshold" = orange,
    "At or above the entry-level threshold" = green
  ), guide = "none") +
  labs(
    title = "Most renter households fall short of the income a starter home now requires",
    subtitle = paste0(
      "Renter households by HUD AMI band, four primary localities<br>",
      "<span style='color:", red, "'>**Below**</span>, ",
      "<span style='color:", orange, "'>**straddling**</span>, and ",
      "<span style='color:", green, "'>**at or above**</span> the income needed to buy an entry-level home"
    ),
    caption = chas_cap("7"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

*Below the entry-level threshold: household income under 80 percent of the region's AMI (the 0-30, 30-50, and 50-80 percent bands) — too low to reach the price a starter home requires. Straddles the threshold: the 80-100 percent AMI band, which contains the actual entry-level income threshold (about `r round(st$ami_pct)` percent of AMI) — some households in this band clear it, others don't. At or above the threshold: the 100 percent and higher AMI band, which fully clears it.*

- **Only about `r label_percent(accuracy = 1)(st$renter_pct_qualifying)` of the region's `r label_comma()(round(st$renter_total_pha))` renter households — roughly `r label_comma()(round(st$renter_qualifying))` — had income at or above the `r label_dollar()(round(st$income_needed, -2))` needed to buy the region's entry-level home over the past 12 months.**
- That leaves roughly `r label_comma()(round(st$renter_total_pha - st$renter_qualifying))` renter households priced out of even the least expensive quarter of homes sold regionally, concentrated in the `r label_comma()(bands$households[bands$income_band == "0-30% AMI"])` households below 30 percent AMI and `r label_comma()(bands$households[bands$income_band == "30-50% AMI"])` more in the 30-50 percent band.
- The entry-level threshold (`r round(st$ami_pct)` percent of the region's 4-person area median income) falls inside the 80-100 percent AMI band, so even moderate-income renter households only partially clear it.

## What renters need to buy their first home

```{r}
#| label: fig-min-income
#| fig-cap: "Even entry-level homes now require six-figure household income across the board"
#| fig-alt: "Column chart showing the share of four-primary-locality households with income at or above what's needed to buy the region's entry-level, typical, and upper-quartile home, using trailing-12-month MLS sale prices and the current 30-year mortgage rate. Entry-level (25th percentile, $335,000) requires about $99,000 and 45 percent of households qualify. Typical (median, $420,000) requires about $123,000 and 35 percent qualify. Upper quartile ($560,000) requires about $162,000 and 23 percent qualify."

income_table |>
  filter(structure_type == "All types") |>
  mutate(price_tier = factor(price_tier, levels = c(
    "Entry-level (25th pct.)", "Typical (median)", "Upper (75th pct.)"
  ))) |>
  ggplot(aes(x = price_tier, y = pct_hh_at_or_above)) +
  geom_col(fill = dblue, width = 0.6) +
  scale_y_continuous(labels = label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Buying even an entry-level home now takes six-figure income",
    subtitle = paste0(
      "Share of households in the four primary localities earning enough to buy<br>",
      "By price point, trailing 12 months of MLS sales at the current mortgage rate"
    ),
    caption = paste0(mls_cap("trailing 12 months"), " Rate uses the current 30-year Freddie Mac PMMS."),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

```{r}
#| label: tbl-min-income
#| tbl-cap: "Minimum income needed to buy, by structure type and price point"

income_table |>
  mutate(
    structure_type = factor(structure_type, levels = c("All types", "Single Family", "Townhouse", "Condominium")),
    price_tier = factor(price_tier, levels = c("Entry-level (25th pct.)", "Typical (median)", "Upper (75th pct.)"))
  ) |>
  arrange(structure_type, price_tier) |>
  transmute(
    `Structure type` = structure_type,
    `Price point` = price_tier,
    `Sale price` = label_dollar()(round(price, -2)),
    `Income needed` = label_dollar()(round(income_needed, -2)),
    `Share of households qualifying` = label_percent(accuracy = 1)(pct_hh_at_or_above)
  ) |>
  kbl() |>
  kable_styling(c("condensed", "striped"))
```

- **Buying the region's median-priced home (`r label_dollar()(round(median_all$price,-2))`) now takes about `r label_dollar()(round(median_all$income_needed,-2))` in household income — only `r label_percent(accuracy = 1)(median_all$pct_hh_at_or_above)` of area households clear that bar.**
- Even the entry-level quartile (`r label_dollar()(round(entry_all$price,-2))`) needs about `r label_dollar()(round(entry_all$income_needed,-2))`, out of reach for more than half of the region's households.
- Condominiums are the most attainable entry point by structure type; single-family entry-level homes need the most income among the three types at this price tier.
- **Assumptions (preliminary):** 10 percent down payment, 30-year fixed term at the current PMMS rate, 1.0 percent of value in annual property tax, and $1,500/year insurance. Not locality-specific. Actual qualifying income will vary with real per-locality tax rates, buyer down payments, and loan products.

<!-- ## Change in housing gap since 2022 -->

<!-- ::: {.callout-note} -->
<!-- ## Since 2022 -->
<!-- The 2022 Framework estimated the region's most common occupations outside business and financial operations could afford homes priced below about `r label_dollar()(base_140k)` (2021 wages, Fig 9.8). -->

<!-- Today's entry-level home price is `r label_dollar()(round(entry_all$price,-2))` — more than double — while the income needed to buy it, `r label_dollar()(round(entry_all$income_needed,-2))`, sits well above what most of those same occupations pay. -->

<!-- The 2022 Framework also put the region's affordable-and-available rental gap at `r label_comma()(base_gap80)` units at 80 percent AMI and `r label_comma()(base_gap30)` units at 30 percent AMI. A comparable current-cycle figure awaits the PUMS-based rebuild of sections 1-2 above; it is not estimated here. -->
<!-- ::: -->
