# Local summaries {#sec-local-summaries}

<!-- Purpose: one stats table per locality — a preliminary scaffold for Phase 7 editing.
     Same metric set for all nine localities; the final edited summaries will curate a
     subset based on local context. No narrative text, no charts. Built in Task 12 pre-work. -->

```{r}
#| label: setup
#| include: false
source("_common.R")
local_stats <- read_rds("data/local_stats.rds")
```

The tables below present a uniform set of key housing indicators for each of the nine localities in the Richmond regional planning area. These are preliminary — the final version will curate a subset of metrics based on local context and add narrative.

**Column notes.** "2022 ref." is the 2022 annual value for PEP (population), BPS (permits), and PSH (HUD units); the 2018–2022 ACS 5-year estimate for income, tenure, home value, rent, and cost burden; and "—" where no comparable prior figure is available. "2024/25" is the 2025 value for PEP and the 2024 value for all other sources. "Change" is percent change for counts and dollar values; percentage-point difference for rates and shares.

```{r}
#| label: tables
#| results: asis

loc_order <- unique(local_stats$geoid)

for (geo in loc_order) {
  loc <- local_stats |> filter(geoid == geo)
  nm  <- unique(loc$short_name)
  grp <- unique(loc$group)

  grp_label <- switch(grp,
    primary   = " *(primary)*",
    secondary = " *(secondary)*",
    ashland   = " *(town)*",
    ""
  )

  cat("\n\n##", nm, grp_label, "\n\n")

  tbl <- loc |>
    select(
      `Metric`      = metric_label,
      `2022 ref.`   = value_2022_fmt,
      `2024/25`     = value_current_fmt,
      `Change`      = change_fmt
    ) |>
    mutate(across(where(is.character), \(x) replace_na(x, "—")))

  print(
    kbl(tbl, format = "html", align = c("l", "r", "r", "r")) |>
      kable_styling(c("condensed", "striped"), full_width = FALSE)
  )

  cat("\n\n")
}
```
