# Homeownership market {#sec-ownership}

<!-- Purpose: for-sale market conditions and homeownership access — sales prices, listings,
     production, homeownership rate by race/ethnicity (B25003A-I). Theme-based takeaway H2s;
     static ggplot only; "Since 2022" callout per section; fig-alt throughout. Built in Task 8.
     Narrative rule: takeaway titles + bullet findings + callouts only — no drafted prose. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

tenure      <- read_rds("data/acs_tenure.rds")
race_group  <- read_rds("data/acs_tenure_race_group.rds")
value_rent  <- read_rds("data/acs_value_rent.rds")
mls         <- read_rds("data/mls_monthly.rds")
bps         <- read_rds("data/bps.rds")
fred        <- read_rds("data/fred.rds")

# Palette anchors — match ggtext subtitle spans to point/line colors
lblue <- pha_pal[2]; dblue <- pha_pal[6]; green <- pha_pal[1]; orange <- pha_pal[3]

clean_name <- function(x) str_remove(x, " County, Virginia| city, Virginia|, Virginia")

# Inline scalars for bullet text
owner_share_24 <- tenure |>
  filter(geoid %in% pha, year == 2024, tenure != "Total") |>
  summarise(est = sum(estimate), .by = tenure) |>
  mutate(share = est / sum(est)) |>
  filter(tenure == "Owner") |>
  pull(share)

# Homeownership by race and ethnicity. r/acs_tenure_race.R does the pooling and
# the standard-error math, including the "region4" row that covers the 4 primary
# localities; this chapter only filters and plots.
race_region <- race_group |> filter(geoid == "region4")

race_rate_24 <- race_region |> filter(year == 2024, tenure == "Owner")

race_rate <- function(r) race_rate_24$owner_rate[race_rate_24$race_group == r]

# Region-wide owner share for the same 4 localities, as the figure's reference.
region_rate_24 <- tenure |>
  filter(geoid %in% pha, year == 2024) |>
  summarise(hh = sum(estimate), .by = tenure) |>
  summarise(rate = hh[tenure == "Owner"] / hh[tenure == "Total"]) |>
  pull(rate)

sale <- mls |>
  filter(metric == "Median sale price") |>
  arrange(month) |>
  mutate(roll12 = as.numeric(stats::filter(price, rep(1 / 12, 12), sides = 1)))

mls_latest <- sale |> slice_max(month, n = 1) |> pull(price)
mls_first  <- sale |> slice_min(month, n = 1) |> pull(price)
```

## Homeownership rates remain stable

```{r}
#| label: fig-owner-share
#| fig-cap: "No significant changes to homeownership rates"
#| fig-alt: "Dumbbell chart comparing the owner-occupied share of households in 2019 and 2024 for the four primary Richmond region localities and Virginia. Hanover dipped slightly, the only primary locality where the owner share fell since 2019. Richmond city is lowest at about 44 percent, the only majority-renter locality."

own <- tenure |>
  filter(geoid %in% c(pha, "51"), tenure == "Owner", year %in% c(2019, 2024)) |>
  mutate(
    locality = clean_name(name),
    locality = fct_reorder(locality, share, .fun = max)
  )

own_wide <- own |>
  select(locality, year, share) |>
  pivot_wider(names_from = year, values_from = share, names_prefix = "y")

ggplot(own, aes(x = share, y = locality)) +
  geom_segment(
    data = own_wide,
    aes(x = y2019, xend = y2024, y = locality, yend = locality),
    color = "grey70", linewidth = 1
  ) +
  geom_point(aes(color = factor(year)), size = 3) +
  scale_color_manual(values = c("2019" = lblue, "2024" = dblue), guide = "none") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "No significant changes to homeownership rates",
    subtitle = paste0(
      "Owner-occupied share of households, ",
      "<span style='color:", lblue, "'>**2019**</span> vs ",
      "<span style='color:", dblue, "'>**2024**</span>"
    ),
    caption = acs_cap("B25003", year = "2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha()
```

- **The region as a whole is `r label_percent(accuracy = 0.1)(owner_share_24)` owner-occupied in 2024, just below the statewide rate of 67.3 percent.**
- Hanover is the only primary locality where the owner share fell since 2019, dipping slightly from 82.5 to 82.1 percent.
- The four core localities barely moved since 2019 — Richmond city (43.5 percent) remains the region's only majority-renter jurisdiction, and Hanover (82.1 percent) the highest among them.

## Racial homeownership gap remains high across region

```{r}
#| label: fig-owner-race
#| fig-cap: "Racial homeownership gap remains high across region"
#| fig-alt: "Horizontal bar chart of the 2024 homeownership rate by race and ethnicity of householder for the four primary Richmond region localities combined, ordered from highest to lowest. White, non-Hispanic households are highest at 74.6 percent and Asian households at 70.7 percent, both above the region-wide rate of 65.1 percent. Multiracial households are at 54.9 percent, Black at 49.8 percent, Hispanic or Latino at 48.3 percent, and another race lowest at 41.7 percent."

race_plot <- race_rate_24 |>
  filter(rate_reliability != "Low") |>
  mutate(race_group = fct_reorder(as.character(race_group), owner_rate))

ggplot(race_plot, aes(x = owner_rate, y = race_group)) +
  geom_col(fill = dblue, width = 0.7) +
  #geom_vline(xintercept = region_rate_24, color = "grey40", linetype = "22") +
  geom_text(aes(label = label_percent(accuracy = 0.1)(owner_rate)),
            hjust = -0.15, size = 3.2, color = "grey20") +
  scale_x_continuous(labels = label_percent(accuracy = 1),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Racial homeownership gap remains high across region",
    subtitle = "Owner-occupied share of households by race and ethnicity of householder", 
    caption = acs_cap("B25003B-I"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("x")
```

- **White, non-Hispanic households own at `r label_percent(accuracy = 0.1)(race_rate("White, non-Hispanic"))`, `r label_number(accuracy = 0.1)(100 * (race_rate("White, non-Hispanic") - race_rate("Black")))` percentage points above Black households at `r label_percent(accuracy = 0.1)(race_rate("Black"))` and `r label_number(accuracy = 0.1)(100 * (race_rate("White, non-Hispanic") - race_rate("Hispanic or Latino")))` points above Hispanic or Latino households at `r label_percent(accuracy = 0.1)(race_rate("Hispanic or Latino"))`.**
- Two groups sit above the region-wide rate of `r label_percent(accuracy = 0.1)(region_rate_24)`: White, non-Hispanic and Asian (`r label_percent(accuracy = 0.1)(race_rate("Asian"))`).
- Another race is the lowest group at `r label_percent(accuracy = 0.1)(race_rate("Another race"))`, and multiracial households sit between the two poles at `r label_percent(accuracy = 0.1)(race_rate("Multiracial"))`.
- Black and Hispanic or Latino households are the region's two largest groups after White, non-Hispanic — `r label_comma()(race_rate_24$total_hh[race_rate_24$race_group == "Black"])` and `r label_comma()(race_rate_24$total_hh[race_rate_24$race_group == "Hispanic or Latino"])` households — so the two lowest large-group rates cover roughly a third of all households in the four localities.

```{r}
#| label: fig-owner-race-local
#| fig-cap: "The ownership gap by race holds in each locality"
#| fig-alt: "Four small horizontal bar charts, one per primary locality, of the 2024 homeownership rate by race and ethnicity of householder. White, non-Hispanic households own at 84.2 percent in Hanover, 83.9 in Chesterfield, 74.0 in Henrico, and 52.6 in Richmond city. Black households trail White, non-Hispanic households by 23.2 points in Henrico, 20.6 in Chesterfield, 16.4 in Richmond city, and 15.3 in Hanover. Richmond city posts the lowest rate for every group, from 52.6 percent for White, non-Hispanic households down to 20.3 percent for another race. Two Hanover cells, Asian and another race, are suppressed for low reliability."

# Two of the 24 locality cells fail CV-30, both in Hanover (Asian renter count
# CV 65.4, another-race renter count CV 32.3). They are dropped, not the whole
# group -- suppression is per cell, per the reliability policy.
race_local <- race_group |>
  filter(geoid %in% pha, year == 2024, tenure == "Owner",
         rate_reliability != "Low") |>
  mutate(
    locality   = clean_name(name),
    race_group = fct_rev(race_group)
  )

ggplot(race_local, aes(x = owner_rate, y = fct_wrap(race_group, 20))) +
  geom_col(fill = dblue, width = 0.7) +
  geom_text(aes(label = label_percent(accuracy = 1)(owner_rate)),
            hjust = -0.2, size = 2.8, color = "grey20") +
  facet_wrap(~ locality, nrow = 2) +
  scale_x_continuous(labels = label_percent(accuracy = 1),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "The ownership gap by race holds in each locality",
    subtitle = paste0(
      "Owner-occupied share of households by race and ethnicity of householder"
    ),
    caption = acs_cap("B25003B-I"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("x")
```

- **The White-Black ownership gap is present in all four primary localities: 23.2 points in Henrico, 20.6 in Chesterfield, 16.4 in Richmond city, and 15.3 in Hanover.**
- Richmond city posts the lowest rate for every group, from 52.6 percent for White, non-Hispanic households to 20.3 percent for another race.
- Asian households have the widest locality spread of any reported group — 86.1 percent in Chesterfield and 68.8 in Henrico against 40.3 percent in Richmond city, a 46-point range.
- Hispanic or Latino households own at 65.4 percent in Hanover and 59.2 in Chesterfield, but 29.9 percent in Richmond city.
- Rates for two groups in Hanover (Asian and Another race) have low data reliability due to limited number of households

## Home prices surged, then started to cool

```{r}
#| label: fig-mls-price
#| fig-cap: "Home prices are up 65 percent since the start of 2020"
#| fig-alt: "Line chart of the PHA region's monthly median home sale price from January 2020 through July 2026 in nominal dollars, with a 12-month rolling average and a shaded band marking 2022 onward. The trend climbs from about 264,000 dollars in January 2020 to roughly 435,000 dollars by mid-2026, with continued gains after 2022 though at roughly half the pace of 2020-2022."

sale_recent <- sale |> filter(month >= as.Date("2020-01-01"))
since_2022  <- sale_recent |> filter(month == as.Date("2022-01-01"))
mls_2020    <- sale_recent |> slice_min(month, n = 1)
mls_latest_r <- sale_recent |> slice_max(month, n = 1)

ggplot(sale_recent, aes(x = month)) +
  annotate("rect", xmin = as.Date("2022-01-01"), xmax = max(sale_recent$month),
           ymin = 0, ymax = Inf, fill = green, alpha = 0.06) +
  annotate("text", x = as.Date("2022-02-01"), y = 60000, label = "Since 2022",
           color = green, fontface = "bold", hjust = 0, size = 3.2) +
  geom_line(aes(y = price), color = lblue, linewidth = 0.4, alpha = 0.8) +
  geom_line(aes(y = roll12), color = green, linewidth = 1.1, na.rm = TRUE) +
  scale_y_continuous(labels = label_dollar(), limits = c(0, NA)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Home prices are up 65 percent since the start of 2020",
    subtitle = paste0(
      "PHA region <span style='color:", lblue, "'>**monthly median sale price**</span> and ",
      "<span style='color:", green, "'>**12-month average**</span>, nominal dollars<br>",
      "January 2020 through July 2026 · shaded band marks the period since 2022"
    ),
    caption = mls_cap("2020-2026"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **The median sale price reached `r label_dollar()(mls_latest_r$price)` in July 2026, up `r label_percent(accuracy = 1)(mls_latest_r$price / mls_2020$price - 1)` from `r label_dollar()(mls_2020$price)` in January 2020.**
- Prices kept climbing after 2022: the 12-month average is up `r label_percent(accuracy = 1)(mls_latest_r$roll12 / since_2022$roll12 - 1)` since January 2022, from about `r label_dollar()(round(since_2022$roll12))` to `r label_dollar()(round(mls_latest_r$roll12))`.
- The steepest appreciation came in 2020-2022, when the 12-month average rose from roughly `r label_dollar()(round(mls_2020$roll12))` to `r label_dollar()(round(since_2022$roll12))` — growth has continued since, but at roughly half that pace.
- Monthly medians are seasonal and noisy; the 12-month average, shown here since 2020, is the better read on trend.

## Home values on the rise to reflect market

```{r}
#| label: fig-home-value
#| fig-cap: "Home values are up at least a third in every locality since 2019"
#| fig-alt: "Horizontal grouped bar chart of median owner-occupied home values in 2019 and 2024 for the four primary Richmond region localities and Virginia, in nominal dollars. Every locality shows a large increase since 2019, with Richmond city rising the most in percentage terms, from 230,500 to 353,000 dollars. Hanover is the highest primary locality in 2024."

val <- value_rent |>
  filter(geoid %in% c(pha, "51"), measure == "Median home value", year %in% c(2019, 2024)) |>
  mutate(
    locality = clean_name(name),
    locality = fct_reorder(locality, estimate, .fun = max)
  )

ggplot(val, aes(x = estimate, y = locality, fill = factor(year))) +
  geom_col(position = position_dodge2(reverse = TRUE), width = 0.7) +
  scale_fill_manual(values = c("2019" = lblue, "2024" = dblue), guide = "none") +
  scale_x_continuous(labels = label_dollar(), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Home values are up at least a third in every locality since 2019",
    subtitle = paste0(
      "Median value of owner-occupied homes, four primary localities and Virginia, ",
      "<span style='color:", lblue, "'>**2019**</span> vs ",
      "<span style='color:", dblue, "'>**2024**</span>, nominal dollars"
    ),
    caption = acs_cap("B25077", year = "2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("x")
```

- **Richmond city posted the fastest home value growth among the four primary localities since 2019 — up 53 percent, from $230,500 to $353,000 — continuing to erode its historical role as the region's lower-cost ownership market.**
- Every primary locality gained at least 33 percent since 2019, and all four now exceed the $350,000 mark.
- Hanover is the only primary locality above the statewide median home value of $383,700; Chesterfield and Henrico sit between $350,000 and the statewide median.
- Medians are locality-level ACS estimates and do not sum or average to a regional figure; all plotted values are High reliability.

## Multifamily permits beginning to outpace single-family

```{r}
#| label: fig-bps-permits
#| fig-cap: "New residential permits shifting away from single-family"
#| fig-alt: "Overlapping line chart of annual building permits by structure type (single-family and multifamily) for the Richmond region, 2000 through 2025, with 2000-2019 shown in muted grey and 2020-2025 shown in color. Single-family permits fell from about 4,825 units in 2020 to 3,792 in 2025. Multifamily permits rose from about 2,770 units in 2020 to 4,677 in 2025, exceeding single-family in 2022, 2023, and 2025."

bps_reg <- bps |>
  filter(geoid %in% pha) |>
  summarise(units = sum(units), .by = c(year, structure))

bps_recent <- bps_reg |> filter(year >= 2020)

sf_2020 <- bps_reg |> filter(structure == "Single-family", year == 2020) |> pull(units)
sf_2025 <- bps_reg |> filter(structure == "Single-family", year == max(year)) |> pull(units)
mf_2020 <- bps_reg |> filter(structure == "Multifamily", year == 2020) |> pull(units)
mf_2025 <- bps_reg |> filter(structure == "Multifamily", year == max(year)) |> pull(units)
bps_latest_yr <- max(bps_reg$year)

ggplot(bps_reg, aes(x = year, y = units, group = structure)) +
  geom_line(color = "grey70", linewidth = 0.7) +
  geom_line(data = bps_recent, aes(color = structure), linewidth = 1.1) +
  scale_color_manual(values = c("Single-family" = dblue, "Multifamily" = orange), guide = "none") +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "New residential permits shifting away from single-family",
    subtitle = paste0(
      "Annual building permits, four primary localities<br>",
      "2000-2019 shown muted for context · ",
      "<span style='color:", dblue, "'>**single-family**</span> and ",
      "<span style='color:", orange, "'>**multifamily**</span> emphasized since 2020"
    ),
    caption = bps_cap("2000-2025"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **Since 2020, multifamily permits are up `r label_percent(accuracy = 1)(mf_2025/mf_2020 - 1)`, from `r label_comma()(mf_2020)` to `r label_comma()(mf_2025)` units, while single-family permits fell `r label_percent(accuracy = 1)(1 - sf_2025/sf_2020)`, from `r label_comma()(sf_2020)` to `r label_comma()(sf_2025)`.**
- 2022 was the first year in the 2000-2025 record that regional multifamily volume exceeded single-family (5,804 vs 3,946 units) — a pattern that repeated in 2023 and again in `r bps_latest_yr`.
- Total regional permitting peaked in 2022 at 9,750 units, more than double the post-recession low of 2,186 in 2009, and has eased only modestly since.

## Rising mortgage rates lock many buyers on the sidelines

```{r}
#| label: fig-pmms-price
#| fig-cap: "Mortgage rates have more than doubled since their 2020-2021 low"
#| fig-alt: "Dual-axis line chart of the 30-year fixed mortgage rate and the PHA region's monthly median home sale price, January 2020 through July 2026. The mortgage rate fell to a low of 2.68 percent in December 2020, rose to a peak of 7.62 percent in October 2023, and stands at 6.54 percent in July 2026. The median sale price rose from 264,000 dollars in January 2020 to 435,000 dollars in July 2026, with no pause during the rate run-up."

rate_price <- fred |>
  filter(month >= as.Date("2020-01-01")) |>
  inner_join(sale |> select(month, price), by = "month")

scale_factor <- 55000   # maps the rate (%) onto the price-dollar axis for the overlay

rate_lo <- rate_price |> slice_min(rate, n = 1)
rate_hi <- rate_price |> slice_max(rate, n = 1)
rate_latest <- rate_price |> slice_max(month, n = 1)

ggplot(rate_price, aes(x = month)) +
  geom_line(aes(y = price), color = dblue, linewidth = 1.1) +
  geom_line(aes(y = rate * scale_factor), color = orange, linewidth = 1.1) +
  scale_y_continuous(
    labels = label_dollar(), limits = c(0, NA),
    sec.axis = sec_axis(~ . / scale_factor, labels = label_percent(scale = 1))
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Mortgage rates have more than doubled since their 2020-2021 low",
    subtitle = paste0(
      "PHA region <span style='color:", dblue, "'>**median sale price**</span> and ",
      "<span style='color:", orange, "'>**30-year fixed mortgage rate**</span><br>",
      "January 2020 through July 2026, nominal dollars"
    ),
    caption = pmms_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **The 30-year fixed rate fell to a low of `r label_percent(accuracy = 0.01, scale = 1)(rate_lo$rate)` in `r format(rate_lo$month, "%B %Y")`, peaked at `r label_percent(accuracy = 0.01, scale = 1)(rate_hi$rate)` in `r format(rate_hi$month, "%B %Y")`, and stands at `r label_percent(accuracy = 0.01, scale = 1)(rate_latest$rate)` as of `r format(rate_latest$month, "%B %Y")` — more than double the 2020-2021 low.**
- Home prices did not pause for the rate run-up: the median sale price rose from `r label_dollar()(mls_2020$price)` in January 2020 to `r label_dollar()(mls_latest_r$price)` by July 2026, the 65 percent gain covered above.
- Higher rates and higher prices are compounding rather than offsetting — a fixed monthly housing budget buys much less home today than it did when rates sat near 3 percent.
- The mortgage rate is Freddie Mac's national PMMS survey, not Richmond-specific; both series here are nominal, unadjusted for inflation.

<!-- ::: {.callout-note title="To be added"} -->
<!-- **Sales inventory and months of supply.** Active listings, new listings, and months-of-supply trends from the regional MLS. -->
<!-- ::: -->
