# Homeownership market {#sec-ownership}

<!-- Purpose: for-sale market conditions and homeownership access — sales prices, listings,
     production, homeownership rate by race/ethnicity (B25003A-I). Theme-based takeaway H2s;
     static ggplot only; "Since 2022" callout per section; fig-alt throughout. Built in Task 8.
     Narrative rule: takeaway titles + bullet findings + callouts only — no drafted prose. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

tenure     <- read_rds("data/acs_tenure.rds")
value_rent <- read_rds("data/acs_value_rent.rds")
mls        <- read_rds("data/mls_monthly.rds")
bps        <- read_rds("data/bps.rds")
fred       <- read_rds("data/fred.rds")

# Palette anchors — match ggtext subtitle spans to point/line colors
lblue <- pha_pal[2]; dblue <- pha_pal[6]; green <- pha_pal[1]; orange <- pha_pal[3]

clean_name <- function(x) str_remove(x, " County, Virginia| city, Virginia|, Virginia")

# Inline scalars for bullet text
owner_share_24 <- tenure |>
  filter(geoid %in% rr, year == 2024, tenure != "Total") |>
  summarise(est = sum(estimate), .by = tenure) |>
  mutate(share = est / sum(est)) |>
  filter(tenure == "Owner") |>
  pull(share)

sale <- mls |>
  filter(metric == "Median sale price") |>
  arrange(month) |>
  mutate(roll12 = as.numeric(stats::filter(price, rep(1 / 12, 12), sides = 1)))

mls_latest <- sale |> slice_max(month, n = 1) |> pull(price)
mls_first  <- sale |> slice_min(month, n = 1) |> pull(price)
```

## Homeownership rates rose almost everywhere since 2019

```{r}
#| label: fig-owner-share
#| fig-cap: "Homeownership rates rose almost everywhere since 2019"
#| fig-alt: "Dumbbell chart comparing the owner-occupied share of households in 2019 and 2024 for the eight Richmond region localities and Virginia. New Kent rose nearly six percentage points to 93 percent, and Powhatan and Charles City each gained more than two points. Hanover dipped slightly, the only decline. Richmond city is lowest at about 44 percent."

own <- tenure |>
  filter(tenure == "Owner", year %in% c(2019, 2024)) |>
  mutate(
    locality = clean_name(name),
    locality = fct_reorder(locality, share, .fun = max)
  )

own_wide <- own |>
  select(locality, year, share) |>
  pivot_wider(names_from = year, values_from = share, names_prefix = "y")

ggplot(own, aes(x = share, y = locality)) +
  geom_segment(
    data = own_wide,
    aes(x = y2019, xend = y2024, y = locality, yend = locality),
    color = "grey70", linewidth = 1
  ) +
  geom_point(aes(color = factor(year)), size = 3) +
  scale_color_manual(values = c("2019" = lblue, "2024" = dblue), guide = "none") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "Homeownership rates rose almost everywhere since 2019",
    subtitle = paste0(
      "Owner-occupied share of households, Richmond region localities and Virginia<br>",
      "<span style='color:", lblue, "'>**2019**</span> vs ",
      "<span style='color:", dblue, "'>**2024**</span>"
    ),
    caption = acs_cap("B25003", year = "2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha()
```

- **The region as a whole is `r label_percent(accuracy = 0.1)(owner_share_24)` owner-occupied in 2024, just above the statewide rate of 67.3 percent.**
- The three smallest counties posted the largest gains since 2019: New Kent (+6.2 points, to 92.6 percent), Powhatan (+3.1, to 93.2 percent), and Charles City (+2.4, to 86.3 percent).
- Hanover is the only locality where the owner share fell since 2019, dipping slightly from 82.5 to 82.1 percent.
- The four core localities barely moved since 2019 — Richmond city (43.5 percent) remains the region's only majority-renter jurisdiction, and Hanover (82.1 percent) the highest among them.
- All plotted owner-household estimates are High reliability (CV at or below 15); the dataset's only Medium cells are 2024 renter counts in New Kent and Powhatan, and do not appear in this figure.

## Home prices are up 65 percent since the start of 2020

```{r}
#| label: fig-mls-price
#| fig-cap: "Home prices are up 65 percent since the start of 2020"
#| fig-alt: "Line chart of the PHA region's monthly median home sale price from January 2020 through July 2026 in nominal dollars, with a 12-month rolling average and a shaded band marking 2022 onward. The trend climbs from about 264,000 dollars in January 2020 to roughly 435,000 dollars by mid-2026, with continued gains after 2022 though at roughly half the pace of 2020-2022."

sale_recent <- sale |> filter(month >= as.Date("2020-01-01"))
since_2022  <- sale_recent |> filter(month == as.Date("2022-01-01"))
mls_2020    <- sale_recent |> slice_min(month, n = 1)
mls_latest_r <- sale_recent |> slice_max(month, n = 1)

ggplot(sale_recent, aes(x = month)) +
  annotate("rect", xmin = as.Date("2022-01-01"), xmax = max(sale_recent$month),
           ymin = 0, ymax = Inf, fill = green, alpha = 0.06) +
  annotate("text", x = as.Date("2022-02-01"), y = 60000, label = "Since 2022",
           color = green, fontface = "bold", hjust = 0, size = 3.2) +
  geom_line(aes(y = price), color = lblue, linewidth = 0.4, alpha = 0.8) +
  geom_line(aes(y = roll12), color = green, linewidth = 1.1, na.rm = TRUE) +
  scale_y_continuous(labels = label_dollar(), limits = c(0, NA)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Home prices are up 65 percent since the start of 2020",
    subtitle = paste0(
      "PHA region <span style='color:", lblue, "'>**monthly median sale price**</span> and ",
      "<span style='color:", green, "'>**12-month average**</span>, nominal dollars<br>",
      "January 2020 through July 2026 · shaded band marks the period since 2022"
    ),
    caption = mls_cap("2020-2026"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **The median sale price reached `r label_dollar()(mls_latest_r$price)` in July 2026, up `r label_percent(accuracy = 1)(mls_latest_r$price / mls_2020$price - 1)` from `r label_dollar()(mls_2020$price)` in January 2020 (nominal dollars).**
- Prices kept climbing after 2022: the 12-month average is up `r label_percent(accuracy = 1)(mls_latest_r$roll12 / since_2022$roll12 - 1)` since January 2022, from about `r label_dollar()(round(since_2022$roll12))` to `r label_dollar()(round(mls_latest_r$roll12))`.
- The steepest appreciation came in 2020-2022, when the 12-month average rose from roughly `r label_dollar()(round(mls_2020$roll12))` to `r label_dollar()(round(since_2022$roll12))` — growth has continued since, but at roughly half that pace.
- Monthly medians are seasonal and noisy; the 12-month average, shown here since 2020, is the better read on trend.

## Home values are up at least a third in every locality since 2019

```{r}
#| label: fig-home-value
#| fig-cap: "Home values are up at least a third in every locality since 2019"
#| fig-alt: "Horizontal grouped bar chart of median owner-occupied home values in 2019 and 2024 for the eight Richmond region localities and Virginia, in nominal dollars. Goochland is highest in 2024 at 500,600 dollars and Charles City lowest at 241,400 dollars. Every locality shows a large increase since 2019, with Richmond city rising the most in percentage terms, from 230,500 to 353,000 dollars."

val <- value_rent |>
  filter(measure == "Median home value", year %in% c(2019, 2024)) |>
  mutate(
    locality = clean_name(name),
    locality = fct_reorder(locality, estimate, .fun = max)
  )

ggplot(val, aes(x = estimate, y = locality, fill = factor(year))) +
  geom_col(position = position_dodge2(reverse = TRUE), width = 0.7) +
  scale_fill_manual(values = c("2019" = lblue, "2024" = dblue), guide = "none") +
  scale_x_continuous(labels = label_dollar(), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Home values are up at least a third in every locality since 2019",
    subtitle = paste0(
      "Median value of owner-occupied homes, Richmond region localities and Virginia<br>",
      "<span style='color:", lblue, "'>**2019**</span> vs ",
      "<span style='color:", dblue, "'>**2024**</span>, nominal dollars"
    ),
    caption = acs_cap("B25077", year = "2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("x")
```

- **2024 median home values span more than a two-to-one range across the region, from $241,400 in Charles City to $500,600 in Goochland.**
- Richmond city posted the fastest growth since 2019 — up 53 percent, from $230,500 to $353,000 — continuing to erode its historical role as the region's lower-cost ownership market.
- Every locality gained at least 33 percent since 2019, and the four core localities all now exceed the $350,000 mark.
- Four of the eight localities sit above the statewide median of $383,700: Goochland, Powhatan, Hanover, and New Kent.
- Medians are locality-level ACS estimates and do not sum or average to a regional figure; all plotted values are High reliability.

## Multifamily permits have outpaced single-family in three of the last four years

```{r}
#| label: fig-bps-permits
#| fig-cap: "Multifamily permits have outpaced single-family in three of the last four years"
#| fig-alt: "Overlapping line chart of annual building permits by structure type (single-family and multifamily) for the Richmond region, 2000 through 2025, with 2000-2019 shown in muted grey and 2020-2025 shown in color. Single-family permits fell from about 4,825 units in 2020 to 3,792 in 2025. Multifamily permits rose from about 2,770 units in 2020 to 4,677 in 2025, exceeding single-family in 2022, 2023, and 2025."

bps_reg <- bps |>
  filter(geoid %in% rr) |>
  summarise(units = sum(units), .by = c(year, structure))

bps_recent <- bps_reg |> filter(year >= 2020)

sf_2020 <- bps_reg |> filter(structure == "Single-family", year == 2020) |> pull(units)
sf_2025 <- bps_reg |> filter(structure == "Single-family", year == max(year)) |> pull(units)
mf_2020 <- bps_reg |> filter(structure == "Multifamily", year == 2020) |> pull(units)
mf_2025 <- bps_reg |> filter(structure == "Multifamily", year == max(year)) |> pull(units)
bps_latest_yr <- max(bps_reg$year)

ggplot(bps_reg, aes(x = year, y = units, group = structure)) +
  geom_line(color = "grey70", linewidth = 0.7) +
  geom_line(data = bps_recent, aes(color = structure), linewidth = 1.1) +
  scale_color_manual(values = c("Single-family" = dblue, "Multifamily" = orange), guide = "none") +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Multifamily permits have outpaced single-family in three of the last four years",
    subtitle = paste0(
      "Annual building permits, Richmond region (8 localities)<br>",
      "2000-2019 shown muted for context · ",
      "<span style='color:", dblue, "'>**single-family**</span> and ",
      "<span style='color:", orange, "'>**multifamily**</span> emphasized since 2020"
    ),
    caption = bps_cap("2000-2025"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **Since 2020, multifamily permits are up `r label_percent(accuracy = 1)(mf_2025/mf_2020 - 1)`, from `r label_comma()(mf_2020)` to `r label_comma()(mf_2025)` units, while single-family permits fell `r label_percent(accuracy = 1)(1 - sf_2025/sf_2020)`, from `r label_comma()(sf_2020)` to `r label_comma()(sf_2025)`.**
- 2022 was the first year in the 2000-2025 record that regional multifamily volume exceeded single-family (5,804 vs 3,946 units) — a pattern that repeated in 2023 and again in `r bps_latest_yr`.
- Total regional permitting peaked in 2022 at 9,750 units, more than double the post-recession low of 2,186 in 2009, and has eased only modestly since.
- The 2022 Framework's Fig 5.4 baseline for Chesterfield single-family permits — 545 in 2010, 2,202 in 2020 — matches this pull exactly; same metric, same years, no vintage shift.

## Mortgage rates have more than doubled since their 2020-2021 low

```{r}
#| label: fig-pmms-price
#| fig-cap: "Mortgage rates have more than doubled since their 2020-2021 low"
#| fig-alt: "Dual-axis line chart of the 30-year fixed mortgage rate and the PHA region's monthly median home sale price, January 2020 through July 2026. The mortgage rate fell to a low of 2.68 percent in December 2020, rose to a peak of 7.62 percent in October 2023, and stands at 6.54 percent in July 2026. The median sale price rose from 264,000 dollars in January 2020 to 435,000 dollars in July 2026, with no pause during the rate run-up."

rate_price <- fred |>
  filter(month >= as.Date("2020-01-01")) |>
  inner_join(sale |> select(month, price), by = "month")

scale_factor <- 55000   # maps the rate (%) onto the price-dollar axis for the overlay

rate_lo <- rate_price |> slice_min(rate, n = 1)
rate_hi <- rate_price |> slice_max(rate, n = 1)
rate_latest <- rate_price |> slice_max(month, n = 1)

ggplot(rate_price, aes(x = month)) +
  geom_line(aes(y = price), color = dblue, linewidth = 1.1) +
  geom_line(aes(y = rate * scale_factor), color = orange, linewidth = 1.1) +
  scale_y_continuous(
    labels = label_dollar(), limits = c(0, NA),
    sec.axis = sec_axis(~ . / scale_factor, labels = label_percent(scale = 1))
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Mortgage rates have more than doubled since their 2020-2021 low",
    subtitle = paste0(
      "PHA region <span style='color:", dblue, "'>**median sale price**</span> and ",
      "<span style='color:", orange, "'>**30-year fixed mortgage rate**</span><br>",
      "January 2020 through July 2026, nominal dollars"
    ),
    caption = pmms_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **The 30-year fixed rate fell to a low of `r label_percent(accuracy = 0.01, scale = 1)(rate_lo$rate)` in `r format(rate_lo$month, "%B %Y")`, peaked at `r label_percent(accuracy = 0.01, scale = 1)(rate_hi$rate)` in `r format(rate_hi$month, "%B %Y")`, and stands at `r label_percent(accuracy = 0.01, scale = 1)(rate_latest$rate)` as of `r format(rate_latest$month, "%B %Y")` — more than double the 2020-2021 low.**
- Home prices did not pause for the rate run-up: the median sale price rose from `r label_dollar()(mls_2020$price)` in January 2020 to `r label_dollar()(mls_latest_r$price)` by July 2026, the 65 percent gain covered above.
- Higher rates and higher prices are compounding rather than offsetting — a fixed monthly housing budget buys much less home today than it did when rates sat near 3 percent.
- The mortgage rate is Freddie Mac's national PMMS survey, not Richmond-specific; both series here are nominal, unadjusted for inflation.

::: {.callout-note title="To be added"}
**Sales inventory and months of supply.** Active listings, new listings, and months-of-supply trends from the regional MLS.
:::
