# Rental market {#sec-rental}

<!-- Purpose: rental market conditions and the assisted-housing picture — ACS rents and
     renter households, CoStar asking rents and vacancy; NHPD preservation and consolidated
     rental assistance land in Phase 3. Narrative rule: takeaway titles + bullet findings +
     callouts only — no drafted prose. Built in Task 9. -->

```{r}
#| label: setup
#| include: false
source("_common.R")

tenure   <- read_rds("data/acs_tenure.rds")
rents    <- read_rds("data/acs_value_rent.rds") |>
  filter(measure == "Median gross rent")
costar   <- read_rds("data/costar_rental.rds")
nhpd     <- read_rds("data/nhpd.rds")
psh      <- read_rds("data/psh.rds")
baseline <- read_rds("data/baseline_2022.rds")

# Palette anchor for the PSH figure, added alongside the others below
red <- pha_pal[4]

# Short locality labels shared by both ACS frames
short_name <- function(x) str_remove(x, " County, Virginia|, Virginia")

rents  <- rents  |> mutate(locality = short_name(name))
tenure <- tenure |> mutate(locality = short_name(name))

# Palette anchors — subtitle ggtext spans match fills below
lblue <- pha_pal[2]; orange <- pha_pal[3]; dblue <- pha_pal[6]; green <- pha_pal[1]

# Inline scalars for bullets and callouts
rent_va_24   <- rents |> filter(geoid == "51", year == 2024) |> pull(estimate)
rent_hi_24   <- rents |> filter(geoid != "51", year == 2024) |> slice_max(estimate, n = 1)
rent_lo_24   <- rents |> filter(geoid != "51", year == 2024) |> slice_min(estimate, n = 1)
rent_rva     <- rents |> filter(geoid == "51760") |> select(year, estimate)
rva_rent_chg <- with(rent_rva, estimate[year == 2024] / estimate[year == 2014] - 1)

renter_reg <- tenure |>
  filter(tenure == "Renter", geoid %in% rr) |>
  summarise(renters = sum(estimate), .by = year)
reg_add_5yr   <- diff(renter_reg$renters[renter_reg$year %in% c(2019, 2024)])
reg_chg_5yr   <- reg_add_5yr / renter_reg$renters[renter_reg$year == 2019]
reg_add_10yr  <- diff(renter_reg$renters[renter_reg$year %in% c(2014, 2024)])
reg_chg_10yr  <- reg_add_10yr / renter_reg$renters[renter_reg$year == 2014]
rva_add_5yr   <- tenure |>
  filter(geoid == "51760", tenure == "Renter") |>
  summarise(chg = estimate[year == 2024] - estimate[year == 2019]) |> pull(chg)
rva_add_10yr  <- tenure |>
  filter(geoid == "51760", tenure == "Renter") |>
  summarise(chg = estimate[year == 2024] - estimate[year == 2014]) |> pull(chg)
rva_share_of_gain <- rva_add_5yr / reg_add_5yr

costar_reg    <- costar |> filter(locality == "Richmond region")
costar_latest <- costar_reg |> slice_max(quarter, n = 1)
costar_first  <- costar_reg |> slice_min(quarter, n = 1)
costar_chg    <- costar_latest$ask_rent / costar_first$ask_rent - 1
rent_2022q1   <- costar_reg |> filter(period == "2022 Q1") |> pull(ask_rent)

base_peak_rent <- baseline |>
  filter(metric == "avg_asking_rent_regional_peak") |> pull(value)
base_vacancy   <- baseline |>
  filter(metric == "rental_vacancy_rate", geography == "region") |> pull(value)
```

## Median rent has climbed fastest since 2019, in every locality

```{r}
#| label: fig-median-rent
#| fig-cap: "Median rent has climbed fastest since 2019, in every locality"
#| fig-alt: "Grouped horizontal bar chart of median gross rent for eight Richmond-area localities and Virginia in 2014, 2019, and 2024, in nominal dollars. The 2014 bars are muted grey. Every locality's 2024 bar is well above its 2019 bar. New Kent has the highest 2024 median rent at $1,721, up 70 percent since 2019 alone, the sharpest increase in the region. Charles City has the lowest 2024 median rent at $1,099."

rent_order <- rents |> filter(year == 2024) |> arrange(estimate) |> pull(locality)

rents |>
  mutate(
    locality = factor(locality, levels = rent_order),
    year = factor(year)
  ) |>
  ggplot(aes(x = estimate, y = locality, fill = year)) +
  geom_col(position = position_dodge2(reverse = TRUE), width = 0.75) +
  scale_x_continuous(labels = label_dollar(), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c("2014" = "grey80", "2019" = dblue, "2024" = orange),
                    guide = "none") +
  labs(
    title = "Median rent has climbed fastest since 2019, in every locality",
    subtitle = paste0(
      "Median gross rent in nominal dollars, Richmond-area localities and Virginia<br>",
      "2014 shown muted for context · ",
      "<span style='color:", dblue, "'>**2019**</span> to ",
      "<span style='color:", orange, "'>**2024**</span> change is the focus"
    ),
    caption = acs_cap("B25064", year = "2010-2014, 2015-2019, and 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("x")
```

- **Every locality posted a higher median gross rent in 2024 than in 2019, and in all eight the 2019-to-2024 increase outpaced the full 2014-to-2019 span.**
- New Kent had the region's highest 2024 median rent at `r label_dollar()(rent_hi_24$estimate)`, up 70 percent since 2019 alone — the sharpest increase in the region.
- Richmond city's median rent rose 34 percent since 2019, from `r label_dollar()(rent_rva$estimate[rent_rva$year == 2019])` to `r label_dollar()(rent_rva$estimate[rent_rva$year == 2024])` — yet stayed below the statewide median of `r label_dollar()(rent_va_24)`.
- Powhatan had the slowest rent growth since 2019, at 19 percent; small-county medians (New Kent, Powhatan, Goochland, Charles City) carry wide margins of error and describe small renter stocks.
- All figures are nominal dollars; inflation-adjusted comparisons come with the cost-burden analysis.

## Renter shares held steady since 2019, except in the smallest counties

```{r}
#| label: fig-renter-share
#| fig-cap: "Renter shares held steady since 2019, except in the smallest counties"
#| fig-alt: "Dumbbell chart comparing the renter share of households in 2019 and 2024 for eight Richmond-area localities and Virginia. Richmond city has by far the highest renter share, near 57 percent in both years. Henrico is near 36 percent in both years, and Virginia near 33 percent. The six remaining counties are all below 25 percent, and the smallest counties are lowest in 2024: New Kent and Powhatan each fell to about 7 percent."

renter_share <- tenure |>
  filter(tenure == "Renter", year %in% c(2019, 2024)) |>
  mutate(year = factor(year))

share_order <- renter_share |> filter(year == "2024") |>
  arrange(share) |> pull(locality)

renter_share |>
  mutate(locality = factor(locality, levels = share_order)) |>
  ggplot(aes(x = share, y = locality)) +
  geom_line(aes(group = locality), color = "grey70", linewidth = 1) +
  geom_point(aes(color = year), size = 3.5) +
  scale_x_continuous(labels = label_percent(accuracy = 1),
                     expand = expansion(mult = c(0.02, 0.05))) +
  scale_color_manual(values = c("2019" = lblue, "2024" = dblue), guide = "none") +
  labs(
    title = "Renter shares held steady since 2019, except in the smallest counties",
    subtitle = paste0(
      "Renter share of all households, Richmond-area localities and Virginia<br>",
      "<span style='color:", lblue, "'>**2019**</span> vs. ",
      "<span style='color:", dblue, "'>**2024**</span>"
    ),
    caption = acs_cap("B25003", year = "2015-2019 & 2020-2024"),
    x = NULL, y = NULL
  ) +
  theme_pha()
```

- **The region added roughly `r label_comma()(reg_add_5yr)` renter households between 2019 and 2024, a `r label_percent(accuracy = 1)(reg_chg_5yr)` increase — faster growth than the full prior five years.**
- Richmond city drove the gain, adding about `r label_comma()(rva_add_5yr)` renter households since 2019 — roughly `r label_percent(accuracy = 0)(rva_share_of_gain)` of the regional increase — though its renter share held near `r label_percent(accuracy = 1)(tenure |> filter(geoid == "51760", tenure == "Renter", year == 2024) |> pull(share))` because total households grew just as fast.
- Chesterfield and Hanover renter counts grew steadily since 2019, but both counties remain more than three-quarters owner-occupied.
- The three smallest counties — Charles City, New Kent, and Powhatan — each lost renter households since 2019, deepening the ownership tilt of the outer region; New Kent's renter count fell 37 percent.
- Over the full decade, the region added `r label_comma()(reg_add_10yr)` renter households (`r label_percent(accuracy = 1)(reg_chg_10yr)`) — most of that growth came in the last five years.

## Asking rents are up 28 percent since 2020, even as the market has loosened

```{r}
#| label: fig-costar-rent
#| fig-cap: "Asking rents are up 28 percent since 2020, even as the market has loosened"
#| fig-alt: "Line chart of the average multifamily asking rent per unit in the eight-locality Richmond region, quarterly from 2020 through mid-2026, in nominal dollars, with a shaded band marking 2022 onward. The line rises from about $1,235 in early 2020 to $1,422 by early 2022 — the market's tightest point, with vacancy near 5.6 percent — then keeps climbing to $1,583 by the second quarter of 2026, even as vacancy rises to 8.5 percent."

costar_since2020 <- costar_reg |> filter(quarter >= as.Date("2020-01-01"))
costar_2020q1    <- costar_since2020 |> slice_min(quarter, n = 1)
costar_2022q1    <- costar_since2020 |> filter(period == "2022 Q1")
chg_since2020    <- costar_latest$ask_rent / costar_2020q1$ask_rent - 1
chg_since2022    <- costar_latest$ask_rent / costar_2022q1$ask_rent - 1

costar_since2020 |>
  ggplot(aes(x = quarter, y = ask_rent)) +
  annotate("rect", xmin = as.Date("2022-01-01"), xmax = max(costar_since2020$quarter),
           ymin = 0, ymax = Inf, fill = green, alpha = 0.06) +
  annotate("text", x = as.Date("2022-02-15"), y = 1000, label = "Since 2022",
           color = green, fontface = "bold", hjust = 0, size = 3.2) +
  geom_line(color = green, linewidth = 1.1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = label_dollar(), limits = c(950, NA)) +
  labs(
    title = "Asking rents are up 28 percent since 2020, even as the market has loosened",
    subtitle = paste0(
      "Average multifamily <span style='color:", green,
      "'>**asking rent per unit**</span>, nominal dollars<br>",
      "8-locality Richmond region, quarterly, 2020 through 2026 Q2<br>",
      "Shaded band marks the period since 2022"
    ),
    caption = costar_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha()
```

- **Regional average asking rent reached `r label_dollar()(round(costar_latest$ask_rent))` in `r costar_latest$period`, up `r label_percent(accuracy = 1)(chg_since2020)` from `r label_dollar()(round(costar_2020q1$ask_rent))` in early 2020.**
- The market was tightest in early 2022 — vacancy near `r label_percent(accuracy = 1)(costar_2022q1$vac_pct)`, the low point of the period — and rents have risen another `r label_percent(accuracy = 1)(chg_since2022)` since then, to `r label_dollar()(round(costar_latest$ask_rent))`.
- Vacancy has since climbed to `r label_percent(accuracy = 0.1)(costar_latest$vac_pct)` in `r costar_latest$period`, above where it stood at the start of 2020, as new deliveries caught up with and then outpaced demand.
- CoStar coverage is professionally managed multifamily; New Kent and Charles City have only a handful of tracked properties, so locality-level trends there are not meaningful.

::: {.callout-note}
## Since 2022
The 2022 Framework flagged a regional average asking rent of `r label_dollar()(base_peak_rent)` in early 2022 as a two-decade high with vacancy near `r label_percent(accuracy = 1)(base_vacancy / 100)`. Rents have since risen another `r label_percent(accuracy = 1)(costar_latest$ask_rent / rent_2022q1 - 1)` to `r label_dollar()(round(costar_latest$ask_rent))`, but the market has loosened: vacancy is now `r label_percent(accuracy = 0.1)(costar_latest$vac_pct)`, and rent growth has slowed sharply. The 2022 Framework also reported a net *loss* of renter households from 2016 to 2020 (with a COVID-era data caveat); the 2020–2024 ACS instead shows renter households growing across the region.
:::

## Nearly 30 percent of the region's assisted units expire within a decade

```{r}
#| label: fig-nhpd-risk
#| fig-cap: "Nearly 30 percent of the region's assisted units expire within a decade"
#| fig-alt: "Column chart of active federally assisted rental units in the Richmond region by subsidy-expiration window. About 2,142 units expire 2026-2030, 4,349 expire 2031-2035, 3,982 expire 2036-2040, 8,972 expire 2041 or later, and 3,336 carry no listed expiration date, mostly public housing and HOME-assisted properties."

risk_order <- levels(nhpd$expire_window)
risk_pal <- c(
  "2026-2030"             = pha_pal[4],  # Red -- nearest-term risk
  "2031-2035"              = pha_pal[3],  # Orange
  "2036-2040"              = pha_pal[2],  # Light blue
  "2041 and later"         = pha_pal[6],  # Dark blue
  "No listed expiration"   = "grey70"
)

risk <- nhpd |>
  summarise(units = sum(total_units, na.rm = TRUE), .by = expire_window) |>
  mutate(expire_window = factor(expire_window, levels = risk_order))

risk_decade  <- risk |> filter(expire_window %in% c("2026-2030", "2031-2035")) |> pull(units) |> sum()
risk_total   <- sum(risk$units)
risk_share   <- risk_decade / risk_total

ggplot(risk, aes(x = expire_window, y = units, fill = expire_window)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = risk_pal, guide = "none") +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Nearly 30 percent of the region's assisted units expire within a decade",
    subtitle = paste0(
      "Active federally assisted rental units by subsidy-expiration window, rr region<br>",
      "Richmond city, Henrico, Chesterfield, Hanover, and Charles City properties only"
    ),
    caption = nhpd_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **About `r label_comma()(risk_decade)` of the region's `r label_comma()(risk_total)` active assisted units — `r label_percent(accuracy = 1)(risk_share)` — carry subsidies expiring by 2035, the window preservation funders typically prioritize.**
- Three-quarters of tracked properties (152 of 201) are LIHTC-financed, holding 17,465 units; LIHTC's own affordability period can lapse independently of the subsidy end date tracked here, a second layer of risk this figure does not fully capture.
- Richmond city and Henrico account for the large majority of tracked properties (117 and 47 of 201); Goochland, New Kent, and Powhatan have no NHPD-tracked assisted multifamily properties at all.
- About 3,336 units carry no listed subsidy end date, mostly public housing and HOME-assisted properties whose affordability commitments do not lapse the same way project-based subsidies do.

## Housing Choice Vouchers grew as public housing kept shrinking

```{r}
#| label: fig-psh-mix
#| fig-cap: "Housing Choice Vouchers grew as public housing kept shrinking"
#| fig-alt: "Line chart of regional federal rental-assistance units, 2020 through 2024, for two programs. Housing Choice Vouchers rose from 5,910 to 6,822 units. Public housing fell from 3,706 to 3,226 units. Project-based Section 8 held near 5,770 units and is not plotted here."

psh_mix <- psh |>
  filter(program_label %in% c("Housing Choice Vouchers", "Public Housing")) |>
  summarise(units = sum(total_units, na.rm = TRUE), .by = c(year, program_label))

hcv_2020 <- psh_mix |> filter(program_label == "Housing Choice Vouchers", year == 2020) |> pull(units)
hcv_2024 <- psh_mix |> filter(program_label == "Housing Choice Vouchers", year == 2024) |> pull(units)
ph_2020  <- psh_mix |> filter(program_label == "Public Housing", year == 2020) |> pull(units)
ph_2024  <- psh_mix |> filter(program_label == "Public Housing", year == 2024) |> pull(units)

ggplot(psh_mix, aes(x = year, y = units, color = program_label)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Housing Choice Vouchers" = dblue, "Public Housing" = red),
                     guide = "none") +
  scale_x_continuous(breaks = 2020:2024) +
  scale_y_continuous(labels = label_comma(), limits = c(0, NA)) +
  labs(
    title = "Housing Choice Vouchers grew as public housing kept shrinking",
    subtitle = paste0(
      "<span style='color:", dblue, "'>**Housing Choice Vouchers**</span> and ",
      "<span style='color:", red, "'>**public housing**</span> units, rr region, 2020-2024"
    ),
    caption = posh_cap(),
    x = NULL, y = NULL
  ) +
  theme_pha() +
  add_zero_line("y")
```

- **Housing Choice Vouchers grew `r label_percent(accuracy = 1)(hcv_2024/hcv_2020 - 1)` since 2020, from `r label_comma()(hcv_2020)` to `r label_comma()(hcv_2024)` units, while public housing fell `r label_percent(accuracy = 1)(1 - ph_2024/ph_2020)`, from `r label_comma()(ph_2020)` to `r label_comma()(ph_2024)` — a policy-relevant shift from project-based to tenant-based assistance.**
- Regional voucher utilization slipped from 91 percent occupied in 2020 to 87 percent in 2024, even as the voucher supply grew — more vouchers issued, a slightly larger share sitting unused at any given time.
- Project-based Section 8 (5,763 units in 2024) and the small 202/PRAC and 811/PRAC senior- and disability-designated programs (222 and 69 units) have stayed essentially flat since 2020.
- Richmond city and Henrico hold the large majority of the region's assisted units; Charles City, Goochland, New Kent, and Powhatan combined account for fewer than 40 units across all programs. LIHTC production, tracked separately from this HUD administrative extract, remains a future-draft addition.

::: {.callout-note title="Coming in a future draft"}
**Submarket rents and vacancy.** Submarket-level asking rent, vacancy, and new-construction rent detail, extending the regional trend above.
:::
